---
title: "Quiz 2: R basis"
# author: "Type Your Name Here"
# date: today
format: 
    pdf:        
        pdf-engine: lualatex
        template-partials:
            - ../../doc-class.tex
        include-in-header:
            - ../../math.tex
            - text: |
                \AtBeginDocument{
                    \renewcommand{\theexercise}{\arabic{exercise}}
                }
params:
    show_solution: true
filters:
    - ../../solution-filter.lua
crossref:
    chapters: false
---



# Problems

::: {#exr-}
Recall given the radius of a circle $r$, we can calculate the perimeter by using formula $C = 2\pi r$, and the area by using formula $A = \pi r^2$. Assume $r = 3$. Calculate the value of perimeter and area.
:::

::: {.answer}

```{r}
r <- 3
cir <- 2 * pi * r
area <- pi * (r ^ 2)
```

:::


::: {#exr-}
As the natural state, Arkansas has 7 national parks and 52 state parks. How many national or state parks in total at Arkansas? 

- Why the code below doesn’t work?
```{r}
#| eval: false
num_natl_parks = 'seven'
num_state_parks = 52
num_total_parks = num_natl_parks + num_state_parks
```

- Please modify the code above.
:::


::: {.answer}
- `seven` is a string, which cannot be computed directly.
- `num_natl_parks <- 7`.
:::


::: {#exr-}
a. Use function `length()` to find how many values in the vector a defined below?
b. How many values are greater than 1000?
c. What is the percentage of values in the vector greater than 1000?

```{r}
a = c(275, 2168.36, 2931.5, 275.5, 399.93, 2, 545.128, 209.76, 400, 50.1499, 150, 1595.978)
```
:::


::: {#exr-}
Complete the following tasks.

a. Install and load the package `nycflights13`.
b. Load the data set `flights`.
c. Find the variable names of `flights`.
d. Find the dimension of `flights`.
e. Display the first 3 rows of `flights`.
:::


::: {#exr-}
Complete the following tasks.

a. Please download the `titanic_train.csv` dataset and import the data set to R. Save it as `Titanic`.
b. Display the first 5 rows of `Titanic`.
c. Find the average number of siblings / spouses aboard the Titanic (variable name `SibSp`)
d. Find the average of `age`. Why `mean(Titanic$age)` doesn’t work? Please add `na.rm=TRUE` in `mean()` and retry. Please explain the behavior.

:::



