# Regression with qualitative variables


{{< include ../math.qmd >}}

When there are qualitative variables, the whole regression algorithm stay almost the same. The basic idea is still

$$
y=\text{systematic part}+\varepsilon.
$$

The main difference is that a categorical predictor cannot be inserted directly as a number. Instead, we represent group membership by **indicator variables** (also called **dummy variables**), and then fit the model using the same least squares machinery as before.

Suppose a predictor has $g$ categories. After choosing one category as the baseline, we create $g-1$ dummy variables:

$$
x_j=
\begin{cases}
1,& \text{if the observation is in category }j,\\
0,& \text{otherwise.}
\end{cases}
$$

Then the categorical regression model has the form

$$
y=\beta_0+\beta_1x_1+\cdots+\beta_{g-1}x_{g-1}+\varepsilon.
$$

Here:

- $\beta_0$ is the mean response for the baseline group.
- $\beta_j$ is the difference between the expected response of group $j$ and the baseline group.
- $\varepsilon$ is the random error term.

Therefore it is just ordinary linear regression with specially coded predictors.



## A one-way regression model with qualitative variables

Suppose a factor has three groups: A, B, and C. Take A as the baseline, and define

$$
x_1=I(\text{group B}),\quad x_2=I(\text{group C}).
$$

Then the model is

$$
y=\beta_0+\beta_1x_1+\beta_2x_2+\varepsilon.
$$

The fitted means are

$$
\begin{aligned}
E(y\mid A)&=\beta_0,\\
E(y\mid B)&=\beta_0+\beta_1,\\
E(y\mid C)&=\beta_0+\beta_2.
\end{aligned}
$$

Therefore:

- $\beta_0$ is the mean of group A;
- $\beta_1$ is the difference between the mean of group B and the mean of group A;
- $\beta_2$ is the difference between the mean of group C and the mean of group A.

This is the simplest form of categorical regression.








## One-Way ANOVA

One-way ANOVA (analysis of variance) is a classical method for comparing the means of several populations. Suppose we observe a quantitative response variable across $k$ independent groups (or treatments). The primary question is whether the population means are equal:

$$
H_0:\ \mu_1=\mu_2=\cdots=\mu_k
\qquad\text{vs}\qquad
H_a:\ \text{not all }\mu_i\text{ are equal}.
$$

<!-- Rather than comparing group means pairwise, one-way ANOVA assesses whether the variability **between groups** is large relative to the variability **within groups**. If group means differ substantially compared with the variation among observations within each group, this provides evidence against $H_0$. -->

The traditional one-way ANOVA model is

$$
y_{ij}=\mu+\tau_i+\varepsilon_{ij},\qquad i=1,\ldots,k,\ j=1,\ldots,n_i,
$$

where

- $\mu$ is the overall mean,
- $\tau_i$ is the effect of group $i$ (with $\sum_{i=1}^k \tau_i=0$), and hence the mean of group $i$ is $\mu_i=\mu+\tau_i$,
- $\varepsilon_{ij}\overset{iid}{\sim}N(0,\sigma^2)$.



ANOVA partitions the total variability into **between-group** and **within-group** components and uses an $F$-test to determine whether the group means differ significantly.




Suppose we observe

$$
y_{ij},\quad i=1,\dots,k,\ j=1,\dots,n_i,
$$

where:

- $k$ = number of groups  
- $n_i$ = sample size in group $i$  
- $N=\sum n_i$ = total sample size  

Define:

- group means  
    $$
    \bar y_i=\frac{1}{n_i}\sum_{j=1}^{n_i}y_{ij}
    $$
- overall mean  
    $$
    \bar y=\frac{1}{N}\sum_{i=1}^k\sum_{j=1}^{n_i}y_{ij}.
    $$
- Between-group sum of squares: this measures how far group means are from the overall mean.
    $$
    SSB=\sum_{i=1}^k n_i(\bar y_i-\bar y)^2.
    $$
- Between-group variance (mean square): this is called the **between-group variance** in ANOVA.
    $$
    MSB=\frac{SSB}{k-1}.
    $$
- Within-group sum of squares: this measures variation within each group.
    $$
    SSW=\sum_{i=1}^k\sum_{j=1}^{n_i}(y_{ij}-\bar y_i)^2.
    $$
- Within-group variance (mean square): this is called the **within-group variance** and serves as the pooled estimate of $\sigma^2$ under $H_0$.
    $$
    MSW=\frac{SSW}{N-k}.
    $$
- The one-way ANOVA test statistic is the F-test with the F-statistic:
    $$
    F=\frac{MSB}{MSW}.
    $$

All values mentioned above are summarized in the following standard One-Way ANOVA table

| Source | SS | df | MS | F |
|--------|----|----|----|---|
| Between (Groups) | $SSB$ | $k-1$ | $\frac{SSB}{k-1}$ | $\frac{MSB}{MSW}$ |
| Within (Error) | $SSW$ | $N-k$ | $\frac{SSW}{N-k}$ |  |
| Total | $SST$ | $N-1$ |  |  |

Similar to all the ANOVA tables we mentioned before, we have a decomposition of variance that $SST=SSB+SSW$. 



::: {#exm-}

Scores:

- A: 78, 82, 85, 80  
- B: 88, 90, 85, 87  
- C: 72, 75, 70, 74  

Group means:

$$
\bar y_A=81.25,\quad 
\bar y_B=87.5,\quad 
\bar y_C=72.75,\quad 
\bar y=80.5.
$$

$$
\begin{split}
SSB&=4(81.25-80.5)^2+4(87.5-80.5)^2+4(72.75-80.5)^2=434,\\
MSB&=\frac{434}{2}=217,\\
SSW&=26.75+13+14.75=54.5,\\
MSW&=\frac{54.5}{9}=6.06.
\end{split}
$$
Then 
$$
F=\frac{MSB}{MSW}=\frac{217}{6.06}\approx 35.81.
$$
Under the null hypothesis $H_0:\mu_A=\mu_B=\mu_C$, we have

$$
F\sim F_{2,9}.
$$

So the p-value is

$$
p=P(F_{2,9}\ge 35.81)\approx 5.4\times 10^{-5}.
$$

Since the p-value is much smaller than $0.05$, we reject $H_0$ and conclude that not all group means are equal.


:::





## Relation between the two models

One-way ANOVA can be viewed as a special case of multiple linear regression in which the predictor is categorical. Suppose the explanatory variable has $k$ groups. Then we have the regression model


$$
y=\beta_0+\beta_1x_1+\beta_2x_2+\cdots+\beta_{k-1}x_{k-1}+\varepsilon,
$$

where each $x_j$ is a dummy variable indicating membership in one of the groups, and one group is treated as the reference group.

In this formulation:

- $\beta_0$ is the mean of the reference group,
- each $\beta_j$ measures the difference between the mean of group $j$ and the mean of the reference group.


Thus, testing whether all group means are equal is equivalent to testing
$$
H_0:\beta_1=\beta_2=\cdots=\beta_{k-1}=0.
$$

So the usual one-way ANOVA $F$-test is exactly the same as the overall $F$-test in the corresponding regression model with a categorical predictor.

In this sense, one-way ANOVA and one-way categorical regression are not different methods; they are simply two equivalent ways to describe the same model. ANOVA emphasizes partitioning variation into between-group and within-group components, while regression emphasizes coefficients and predictor effects.




::: {.callout-tip}
## History of ANOVA and regression
Analysis of variance (ANOVA) and regression share a common origin in early 20th-century statistics. Regression was developed earlier by Francis Galton and later formalized by Karl Pearson in studying relationships between quantitative variables. ANOVA was introduced in the 1920s by Ronald A. Fisher to analyze experimental data by decomposing variability into systematic and random components. By the mid-20th century, especially through the work of C. R. Rao and others, ANOVA was recognized as a special case of the general linear model, showing that both methods are unified within the same mathematical framework.
:::




::: {.callout-note}
##  Relation between ANOVA tables
Sicne the independent variable is dummy variable which indicates the group index, the regreesion model simply means that the response variable is modeled by the group mean. In other words, once we decide to talk about the i-th group, the fitted value is the group mean $\hat y_{ij}=\bar y_i$. 

Then 
$$
\begin{split}
SSR&=\sum_{i=1}^k\sum_{j}^{n_i}(\hat y_{ij}-\bar y)^2=\sum_{i=1}^k\sum_{j}^{n_i}(\bar y_i-\bar y)^2=\sum_{i=1}^kn_i(\bar y_i-\bar y)^2=SSB,\\
SSE&=\sum_{i=1}^k\sum_{j}^{n_i}(y_{ij}-\hat y_{ij})^2=\sum_{i=1}^k\sum_{j}^{n_i}(y_{ij}-\bar y_{i})^2=SSW.
\end{split}
$$

Therefore the One-Way ANOVA table is exactly the ANOVA table for this particular multiple linear regression model.

:::




::: {.callout-caution}

## Degree of freedom of predictors

The predictor is also called a factor, and the number of groups is the levels of the factor. Just like mentioned previously, the factor can be modeled by dummy variables. However, since there is only one factor, we treat it as a single factor variable with degree of freedom $k-1$.

Therefore although the regression form seems to have $k-1$ variables, it is essentially a regression model with one variable of degree of freedom of $k-1$.


:::

::: {.callout-note}
## Type III ANOVA table and the summary table

The summary table reports individual t-tests for each coefficient, while the Type III ANOVA table reports a nested-model F-test for each model term. They answer the same hypothesis only when the term has **one degree of freedom**. In that case,
$$
F=t^2,
$$

so the Type III test and the corresponding row in the summary table give identical results.

However, for qualitative variables with multiple levels, the term typically has more than one degree of freedom. In that case:

- the summary table tests coefficients individually (e.g., each group vs the reference),
- the Type III ANOVA table tests the term jointly.

So they generally answer different questions and may give different conclusions.
:::







## R code

In R, we use `factor` to deal with qualitative variables. 


```{r}
x <- factor(c(1,2,3,1,2,3,1,1,2))
x
```

When using `lm` to process a linear regression model, if a variable is a factor, it will be automatically changed into a dummy variable. 


::: {#exm-}

We could genearte a data of 3 groups.

```{r}
n <- 5 
group <- factor(rep(c("A", "B", "C"), each = n))

y <- c(
    rnorm(n, mean = 0, sd = 1),
    rnorm(n, mean = 1, sd = 1),
    rnorm(n, mean = 2, sd = 1)
)

dat <- data.frame(group, y)
dat
```




```{r}
#| eval: false
#| code-fold: true

# This is a function to generate arbitrary groups of data with customized mean and standard deviations.

gen_groups <- function(mu, sd, n = 10, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)

    Lmu <- length(mu)
    Lsd <- length(sd)

    if (Lmu == Lsd) {
        k <- Lmu
    } else if (Lmu > Lsd && Lmu %% Lsd == 0) {
        sd <- rep(sd, length.out = Lmu)
        k <- Lmu
    } else if (Lsd > Lmu && Lsd %% Lmu == 0) {
        mu <- rep(mu, length.out = Lsd)
        k <- Lsd
    } else {
        stop("Lengths of mu and sd must match or be multiples (broadcastable).")
    }

    if (length(n) == 1) n <- rep(n, k)

    if (length(n) != k) {
        stop("n must be length 1 or same as number of groups")
    }

    y <- unlist(Map(
        function(m, s, ni) {
            rnorm(ni, mean = m, sd = s)
        },
        mu, sd, n
    ))

    group <- factor(rep(seq_len(k), times = n))

    data.frame(group, y)
}

dat <- gen_groups(c(1, 2, 3), c(2, 1, 1), n=10)

```






```{r}
fit <- lm(y ~ group, data=dat)
summary(fit)
```

R automatically converts the factor variable into dummy variables. The first level of the factor (which is group A in this example) is used as the base group.

We can relevel the factor to change the base group.

```{r}
dat$group <- relevel(dat$group, ref = "B")
fit <- lm(y ~ group, data = dat)
summary(fit)
```


In this case, all three types of ANOVA tables as well as the ANOVA table for the linear regression are the same, because the model contains only one factor variable. They are different from the summary table, because the summary table tests the individual dummy-variable coefficients, while the ANOVA tables test the factor as a whole.

```{r}
anova(fit)
```

```{r}
library(car)
Anova(fit, type = 2)
```

```{r}
Anova(fit, type = 3)
```

```{r}
#| echo: false

source("../../anova_alt.R")
```

```{r}
anova_alt(fit)
```

:::







## Categorical predictors together with quantitative predictors

Categorical variables can be combined with quantitative variables in the same linear model. For example,

$$
y=\beta_0+\beta_1x+\beta_2z_1+\beta_3z_2+\varepsilon,
$$

where $x$ is quantitative and $z_1,z_2$ are dummy variables for a three-level factor.

In this model:

- the factor shifts the intercept across groups;
- the slope of $x$ is the same for all groups.

If we believe the slope should also vary by group, then we add interaction terms:

$$
y=\beta_0+\beta_1x+\beta_2z_1+\beta_3z_2+\beta_4xz_1+\beta_5xz_2+\varepsilon.
$$

Now each group may have a different regression line.



::: {#exm-}

```{r}
set.seed(1)

n <- 40
group <- factor(rep(c("A", "B", "C"), each = n))

x <- rnorm(3 * n)

z1 <- as.numeric(group == "B")
z2 <- as.numeric(group == "C")

y <- 2 + 1.5 * x + 2 * z1 - 1 * z2 + 1 * x * z1 - 0.5 * x * z2 + rnorm(3 * n)

dat <- data.frame(y, x, group)

fit <- lm(y ~ x * group, data = dat)
```


```{r}
summary(fit)
```

```{r}
anova_alt(fit)
```



- Group A is the base group, which is $\Exp(y)=`r coef(fit)['(Intercept)']`+`r coef(fit)['x']`x$. 
- Group B, compared with A, its intercept is increased by `r coef(summary(fit))['groupB', 'Estimate']`, and its slope is increased by `r coef(summary(fit))['x:groupB', 'Estimate']`. Therefore the model for group B is $\Exp(y)=`r coef(fit)['(Intercept)']+coef(summary(fit))['groupB', 'Estimate']`+`r coef(fit)['x']+coef(summary(fit))['x:groupB', 'Estimate']` x$.
- Group C, compared with A, its intercept is increased by `r coef(summary(fit))['groupC', 'Estimate']`, and its slope is increased by `r coef(summary(fit))['x:groupC', 'Estimate']`. Therefore the model for group C is $\Exp(y)= `r coef(fit)['(Intercept)']+coef(summary(fit))['groupC', 'Estimate']`+`r coef(fit)['x']+coef(summary(fit))['x:groupC', 'Estimate']` x$.
- From the summary table, both interaction terms (`x:groupB` and `x:groupC`) have p-values greater than 0.05, so there is insufficient evidence to conclude that the slopes differ across groups. Thus, the main differences among groups are primarily in their intercepts.
- From the ANOVA table, however, the interaction term `x:group` is significant overall, indicating that the slopes are not exactly the same across groups when tested jointly.
- Although neither interaction coefficient is individually significant in the summary table, the ANOVA test shows that the interaction is significant overall. This indicates that the slopes are not identical across groups, even though no single slope difference is strong enough to be detected individually.
:::




## Relation to two-way ANOVA

Suppose there are two categorical predictors, say factor A and factor B. Then the two-way ANOVA model without interaction is

$$
y_{ijk}=\mu+\alpha_i+\beta_j+\varepsilon_{ijk}.
$$

This corresponds to a regression model containing dummy variables for factor A and dummy variables for factor B.

If we include interaction, the model becomes

$$
y_{ijk}=\mu+\alpha_i+\beta_j+(\alpha\beta)_{ij}+\varepsilon_{ijk}.
$$

In regression form, this means:

1. include dummy variables for factor A;
2. include dummy variables for factor B;
3. include products of those dummy variables to represent the interaction.

Thus:

- two-way ANOVA without interaction corresponds to an additive regression model;
- two-way ANOVA with interaction corresponds to a regression model with interaction terms between categorical predictors.


::: {#exm-}
Here is a small two-way example.

```{r}
set.seed(321)

A <- factor(rep(c("Low", "High"), each = 18))
B <- factor(rep(rep(c("Morning", "Evening", "Night"), each = 6), times = 2))

y <- c(
  rnorm(6, 70, 3), rnorm(6, 74, 3), rnorm(6, 72, 3),
  rnorm(6, 78, 3), rnorm(6, 85, 3), rnorm(6, 82, 3)
)

fit.add <- lm(y ~ A + B)
fit.int <- lm(y ~ A * B)

anova(fit.add, fit.int)
```

The nested-model comparison checks whether the interaction terms are needed. If the interaction is not significant, the additive model may be adequate. If the interaction is significant, the effect of one factor depends on the level of the other factor.
:::


## Traditional ANOVA language versus regression language

The two vocabularies are slightly different, but they refer to the same underlying model.

| Traditional ANOVA language | Regression language |
|---|---|
| treatment or factor | categorical predictor |
| level | category |
| main effect | coefficients associated with one factor |
| interaction | product terms |
| compare means | test linear combinations of coefficients |
| between-group variation | explained variation due to factor terms |
| within-group variation | residual variation |


## Summary

- Categorical regression is ordinary linear regression with dummy variables.
- A one-way ANOVA model is a regression model with one categorical predictor.
- A two-way ANOVA model is a regression model with two categorical predictors, possibly with interaction.
- The ANOVA F-tests are nested-model tests on groups of regression coefficients.
- Modern linear model theory unifies regression and ANOVA in one framework.
