

## Example: MLR phase

We continue to look at the Boston Housing price dataset.

```{r}
library(MASS)
data(Boston)
```

Recall that 


| Variable  | Description                                                                 | Type                    |
|-----------|-----------------------------------------------------------------------------|-------------------------|
| `crim`    | Per capita crime rate by town                                                | Numerical, continuous   |
| `zn`      | Proportion of residential land zoned for lots over 25,000 sq.ft.            | Numerical, continuous   |
| `indus`   | Proportion of non-retail business acres per town                             | Numerical, continuous   |
| `chas`    | Charles River dummy variable (1 if tract bounds river; 0 otherwise)          | Categorical, nominal   |
| `nox`     | Nitrogen oxides concentration (parts per 10 million)                        | Numerical, continuous   |
| `rm`      | Average number of rooms per dwelling                                         | Numerical, continuous   |
| `age`     | Proportion of owner-occupied units built prior to 1940                       | Numerical, continuous   |
| `dis`     | Weighted mean distance to five Boston employment centers                     | Numerical, continuous   |
| `rad`     | Index of accessibility to radial highways (larger = better accessibility)   | Categorical, ordinal   |
| `tax`     | Full-value property-tax rate per \$10,000                                    | Numerical, continuous   |
| `ptratio` | Pupil–teacher ratio by town                                                  | Numerical, continuous   |
| `black`   | $1000(B_k - 0.63)^2$, where $B_k$ is the proportion of Black residents   | Numerical, continuous   |
| `lstat`   | Percentage of lower-status population                                       | Numerical, continuous   |
| `medv`    | Median value of owner-occupied homes (in \$1,000s)                          | Numerical, continuous   |


We need to change `chas` and `rad` to be factors.

```{r}
Boston$chas <- factor(Boston$chas)
Boston$rad <- factor(Boston$rad)
```

```{r}
#| echo: false
source("../../anova_alt.R")
```

## With all predictors

We start from the full model with all predictors.

```{r}
fit.full <- lm(medv ~ ., data=Boston)
anova_alt(fit.full)
```

All variables as a group are significant.

Now we consider building a model with part of the predictors. We have 13 independent variables to choose. Should all the predictors be included in the model? Or use only a couple of them? Which ones should be in the model? We will decide by using variable selection procedures, which is discussed in the next unit. 

By using variable selection procedures, we could identify the most important predictors. For now, let's say, we are going to fit models with 5 predictors chosen: `lstat`, `rm`, `dis`, `ptratio`, `chas`.


```{r}
Boston_5_pred <- Boston[, c("lstat", "rm", "dis", "ptratio","chas", "medv")]
attach(Boston_5_pred)
```

With `attach`, we could directly use variables `lstat`, `rm`, etc. without mentioning the dataset. Let us focus on these variables.



<!-- First, we model `medv` ($y$) as a function of the lower status of the populatio `lstat` ($x_1$), average number of rooms `rm` ($x_2$), distance to Boston employment centres `dis` ($x_3$), pupil-teacher ratio `ptratio` ($x_4$), and if tract bounds river `chas` ($x_5$).  -->




## Is `chas` significantly related to `medv`?

To answer this question, we fit the regression model with `medv` as the $y$ and `chas` as the independent variable $x$. Note that, `chas` is a dummy variable with 2 options. Therefore not tract bounds river (`chas=0`) is chosen as the base level. 

```{r}
fit_chas<-lm(medv ~ chas)
summary(fit_chas)
```

```{r}
anova_alt(fit_chas)
```

From the output above, we could find on average, the expected value of the median value of a house is $6346 higher for the houses tract bounds river than those not.

We could also perfor a hypothesis test as follows:

- $H_0:$ `chas` is not a significant predictor.
- $H_a:$ `chas` is a significant predictor.

Since p-value = 0.0000739 $<\alpha=$ 0.05, the null hypothesis is rejected. We conslude `chas` is a significant predictor. However, on the other hand, the coefficient of determination $r^2=0.03$, which is quite small value. In other word, in addition to `chas`, we need more predictors in the model.


## The relationship between `lstat` and `medv`

In Chapter 3, we fitted stright-line model to relate `medv` ($y$) and `lstat` ($x$). 

```{r}
plot(Boston$lstat, Boston$medv)
```

```{r}
fit_lstat <- lm(medv ~ lstat)
```

The scatterplot shows curvilinear relationship, which suggests a curivilinear model might be a better fit. We will fit the curvilinear model $y=\beta_0+\beta_1 x+\beta_2 x^2+\varepsilon$. 

```{r}
fit_quad<-lm(medv ~ lstat+I(lstat^2))
summary(fit_quad)
```

```{r}
anova_alt(fit_quad)
```

Firt, we test the overall significance of this curvilinear model.

- $H_0: \beta_1=\beta_2=0$
- $H_a:$ At least one of the $\beta$'s $\neq 0$

Since the p-value is almost 0, which is less than $\alpha=0.05$, we reject $H_0$ and conclude the curvilinear model is adequate. 

In order to identify whether the curvilinear term `lstat`$^2$ should be kept in the model, we could run the $t$-test, or partial $F$-test. Since there is only one additional factor (i.e. `lstat`$^2$) is considered here, these two tests are equivalent.

```{r}
anova(fit_lstat, fit_quad)
```

$H_0: y=\beta_0+\beta_1 x$ (Reduced Model)

$H_a:y=\beta_0+\beta_1 x +\beta_2 x^2$ (Complete Model)

Since the $p$-value =7.63e-28 $<\alpha$, we reject $H_0$ and conclude the complete model is prefered. Therefore, `medv` and `lstat` are related in a curvilinear way, with upward curvature (as $\beta_2>0$). 

## Interaction between `rm` and `dis`

An analyst hypothesizes that the efect of distance to Boston employment centres on house price will be greater when there are more room numbers. To test this hypothesis, we will need to run an interaction model for $E(y)$ as a function of `rm` ($x_1$), `dis`($x_2$), and `rm:dis` ($x_1x_2$).

```{r}
fit_interaction <- lm(medv ~ rm * dis)
summary(fit_interaction)
```

```{r}
anova_alt(fit_interaction)
```


- $H_0$: There is no significant interaction between `rm` and `dis`.
- $H_a:$ There is a significant interaction between `rm` and `dis`.

From the p-value = 0.0000042 $<\alpha=.05$ for `rm:dis` in the summary table, we conclude there is a significant interaction between `rm` and `dis`. As the distance from Boston employment centres and number of rooms interact positivel on median house price.

