# Example: MLR phase (continue)

In the previous section we talk about our main tools to handle the `Boston` dataset. In this section we display how to explore the dataset from scratch.



| Variable  | Description                                                                 | Type                    |
|-----------|-----------------------------------------------------------------------------|-------------------------|
| `crim`    | Per capita crime rate by town                                                | Numerical, continuous   |
| `zn`      | Proportion of residential land zoned for lots over 25,000 sq.ft.            | Numerical, continuous   |
| `indus`   | Proportion of non-retail business acres per town                             | Numerical, continuous   |
| `chas`    | Charles River dummy variable (1 if tract bounds river; 0 otherwise)          | Categorical, nominal   |
| `nox`     | Nitrogen oxides concentration (parts per 10 million)                        | Numerical, continuous   |
| `rm`      | Average number of rooms per dwelling                                         | Numerical, continuous   |
| `age`     | Proportion of owner-occupied units built prior to 1940                       | Numerical, continuous   |
| `dis`     | Weighted mean distance to five Boston employment centers                     | Numerical, continuous   |
| `rad`     | Index of accessibility to radial highways (larger = better accessibility)   | Categorical, ordinal   |
| `tax`     | Full-value property-tax rate per \$10,000                                    | Numerical, continuous   |
| `ptratio` | Pupil–teacher ratio by town                                                  | Numerical, continuous   |
| `black`   | $1000(B_k - 0.63)^2$, where $B_k$ is the proportion of Black residents   | Numerical, continuous   |
| `lstat`   | Percentage of lower-status population                                       | Numerical, continuous   |
| `medv`    | Median value of owner-occupied homes (in \$1,000s)                          | Numerical, continuous   |



## Phase 1: Understand the Data Before Fitting Anything
Before touching `lm()`, the first step is always exploratory data analysis (EDA). 


```{r}
library(MASS)
data(Boston)

Boston$chas <- factor(Boston$chas)
Boston$rad <- factor(Boston$rad)

summary(Boston)
```


```{r}
pairs(Boston)
```


```{r}
#| warning: false
num_vars <- Boston[, sapply(Boston, is.numeric)]
cor(num_vars)

library(ggcorrplot)
ggcorrplot(cor_matrix, hc.order = TRUE)
```

What I'm looking for:

- Skewed variables (may need log transforms — `medv` and `crim` are classically right-skewed)
- Non-linear relationships between predictors and `medv`
- High pairwise correlations between predictors (multicollinearity warning signs)
- Outliers or influential observations

Here are some key takeaways from the observations:

- `lstat` vs `medv`: curved, which indicates quadratic or log transform likely needed
- `crim` is extremely right-skewed, which indicates `log(crim)` often works better
- `rm` vs `medv`: mostly linear but with some spread
- `nox`, `dis`, `age` are highly intercorrelated (all related to urban density)
- `tax` and `rad` are nearly collinear

## Phase 2: Start with All Predictors — Full Model


```{r}
fit_full <- lm(medv ~ ., data = Boston)
summary(fit_full)
```
Why: This gives me the lay of the land. I check:

- Adjusted $R^2$ (baseline: ~0.73 for raw Boston)
- Which predictors are significant in the presence of all others
- Residual standard error
- Overall F-test

Then immediately do residual diagnostics on this full model:
```{r}
par(mfrow = c(2, 2))
plot(fit_full)
```

The four diagnostic plots give me:

1. Residuals vs Fitted — checks linearity and homoscedasticity
2. Normal Q-Q — checks normality of residuals
3. Scale-Location — another view of heteroscedasticity
4. Residuals vs Leverage — identifies influential points (Cook's distance)

On raw Boston data, these plots typically show a fan shape (heteroscedasticity) and non-linearity — which tells me transformations are needed before I proceed.


## Phase 3: Consider Transformations

### Check if `log(medv)` improves things

I'd also use Box-Cox to formally choose the best power transformation for `medv`:

```{r}
library(MASS)
boxcox(fit_full)  
```

Reading this specific plot

- The peak appears to be right around $\lambda\approx0$, and 
- the 95% CI dotted lines straddle 0 closely. They look to be roughly $(-0.1, 0.2)$ or so.

By the standard rule, if $\lambda=0$ is inside the 95% CI, `log(y)` is the transformation.

```{r}
fit_log <- lm(log(medv) ~ ., data = Boston)
plot(fit_log)
```

```{r}
summary(fit_log)
```

With `log(medv)` as the response:

- Coefficients are now interpreted as percent changes
- The residual fan shape (heteroscedasticity) is slightly improved