


## R code

How to use `lm` function.

```{r}
#| eval: false
x1 <- runif(100)
x2 <- runif(100)
eps <- rnorm(100, 0, 0.5)
y <- 1+2*x1+3*x2+eps
model <- lm(y~x1+x2)
```

1. Model coefficients

```{r}
#| eval: false
model$coefficients
model$coefficients["(Intercept)"]
model$coefficients["x1"]
model$coefficients["x2"]
```

When using SLR, we could directly read the coefficients and plot lines.
```{r}
#| eval: false
plot(x1, y)
abline(lm(y ~ x1)$coefficients)
```


2. Some computed values
   
We could directly find the residuals and the fitted values. 

```{r}
#| eval: false
model$residuals
model$fitted.values
```

3. Model properties
The standard error, $R^2$, $R_a^2$, and $F$-statistic can be gotten from here.

```{r}
#| eval: false
summary(model)$sigma
summary(model)$r.squared
summary(model)$adj.r.squared
summary(model)$fstatistic
```
The individual Std. error and t value and p value can also be read directly.

```{r}
#| eval: false
summary(model)$coefficients
summary(model)$coefficients['x2', 't value']
```




2. Why Does summary(lm()) Show t-tests Instead of F-tests?
Regression focus:

summary(lm()) emphasizes coefficient estimates (slopes), so t-tests are more intuitive.

It answers: "Is this specific predictor’s slope ≠ 0?"

ANOVA focus:

anova(lm()) emphasizes variance explained by each term.

It answers: "Does this predictor (or model) explain significant variance?"



4. Key Takeaways
For simple regression (1 predictor):

The t-test (from summary) and F-test (from anova) are identical.

For multiple predictors:

t-tests assess individual slopes (adjusting for other predictors).

F-tests assess groups of predictors (e.g., all levels of a categorical variable).

ANOVA is a generalization:

t-tests are a special case of ANOVA when testing a single predictor.