---
title: "HW 3"
# author: "Type Your Name Here"
# date: today
format: 
    pdf:
        pdf-engine: lualatex
        template-partials:
            - ../../doc-class.tex
        include-in-header:
        - text: |
            \AtBeginDocument{
                \renewcommand{\theexercise}{\arabic{exercise}}
            }
params:
    show_solution: true
filters:
    - ../../solution-filter.lua
crossref:
    chapters: false
---






::: {#exr-}
## Dataset `B.csv`

Variables:

- $y$: scale ranging from 8 to 40
- $x_1$: scale ranging from 1 to 10
- $x_2$: scale ranging from 2 to 6
- $x_3$: scale ranging from 8 to 18
- $x_4$: 1 or 0 
  

```{r}
#| echo: false
df <- read.csv('datasets/B.csv')
```



(a) Fit the first-order model, $E(y)=\beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_3+\beta_4x_4$, to the data in the file. Give the least squares prediction equation
(b) Interpret the $\beta$ estimates in the words of the problem. You need to find the values of $\hat\beta_0, \hat\beta_1, \hat\beta_2, \hat\beta_3, \hat\beta_4$ and interpret them one by one.
(c) Is the overall model statistically useful for predicting $y$? Test using $\alpha=.01$. (When performing the hypothesis testing, do not forget to write down the hypotheses first!)
(d) Which statistic, $R^2$ or $R^2_a$, is the preferred measure of model fit? Practically interpret the value of this statistic.
(e) Conduct a test to determine whether $y$ to have decreases linearly as $x_2$. Use $\alpha=0.05$.
(f) Find a $95\%$ confidence interval for $\beta_4$. Practically interpret the result.
(g) Assuming there is a data with value $x_1=7$, $x_2=4$, $x_3=14$,$x_4=1$ Interpret the confidence interval for $E(y)$ for it.

:::




::: {.answer}


```{r}
dat <- read.csv("datasets/B.csv")
dat$x4 <- factor(dat$x4)
fit <- lm(y ~ ., data = dat)
```



```{r}
summary(fit)
```


```{r}
confint(fit)
```

```{r}
newx <- data.frame(x1 = 7, x2 = 4, x3 = 14, x4 = 1)
newx$x4 <- factor(newx$x4, levels=c(0,1))
predict(fit, newx, interval = 'confidence')
```

:::




