---
title: "HW 8"
# author: "Type Your Name Here"
# date: today
format: 
    pdf:
        pdf-engine: lualatex
        template-partials:
            - ../../doc-class.tex
        include-in-header:
        - ../../math.tex
        - text: |
            \AtBeginDocument{
                \renewcommand{\theexercise}{\arabic{exercise}}
            }
params:
    show_solution: false
filters:
    - ../../solution-filter.lua
crossref:
    chapters: false
---


::: {#exr-}
## Data set: `A.csv`
Fit a first-order model for $y$ as a function of $x_1$, $x_2$ and $x_3$.

Check the assumption of a constant error variance by plotting the model residuals against predicted arsenic level. Interpret the results.
:::


::: {.answer}

```{r}
dat <- read.csv("A.csv")
fit <- lm(y~x1+x2+x3, data=dat)
plot(fit, which=1)
```
From the residual plots below, it appears that the spread of the residuals increases as the fitted values increase. This indicates the variances may not be constant.
:::

::: {#exr-}
## Data set: `G.csv`
Consider the interaction model $\Exp(y) = \beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2$. 

a. Fit the model to the data, then plot the residuals against predicted $y$. 
b. Is the assumption of a constant error variance reasonably satisfied? If not, suggest how to modify the model.
c. Use a residual graph to check the assumption of normal errors for the interaction model for $y$. Is the normality assumption reasonably satisfied? If not, suggest how to modify the model.
d. Identify any outliers in the interaction model for $y$. Are any of these outliers influential data points? If so, what are your recommendations?

:::


::: {.answer}

```{r}
dat <- read.csv("G.csv")
fit <- lm(y~x1*x2, data=dat)
# plot(fitted.values(fit), resid(fit))
plot(fit, which=1)
```
There is no indication that the assumption of constant variance is violated.


```{r}
# qqnorm(resid(fit))
# qqline(resid(fit))
plot(fit, which=2)
```
As seen in the graphs above, the normality assumption does not appear to be satisfied. The normal probability plot is not a straight line. There appears to be at least one outlier. We may have to use a normalizing transformation of the data or remove the outliers.


```{r}
plot(fit, which=3)
```
```{r}
plot(fit, which=5)
```
There are 3 outliers: 11, 36 and 64. All have the studentized deleted residual (or equivalently, standardized residual) greater than 3. A “high” value of leverage is given by $h_i > \frac{2(k+1)}{n}=\frac{2\cdot(3+1)}{67}\approx0.119$. Analysis of these outliers indicates that Obs. 64 is influential, because leverage value is high and Cooks Distance is greater than 1. Possibly delete these or see if the model should be revised.

:::