---
title: "HW 6"
# author: "Type Your Name Here"
# date: today
format: 
    pdf:
        pdf-engine: lualatex
        template-partials:
            - ../../doc-class.tex
        include-in-header:
        - text: |
            \AtBeginDocument{
                \renewcommand{\theexercise}{\arabic{exercise}}
            }
params:
    show_solution: false
filters:
    - ../../solution-filter.lua
crossref:
    chapters: false
---


::: {#exr-}
We are estimating the relative error $y$ with a dataset of $n = 49$ observations. 8 independent binary variables were evaluated as potential predictors, which are all formulated as dummy variables. Please answer the following questions. All these questions are abstract questions that no real dataset is provided.

a. In Step 1 of the stepwise regression, how many different one-variable models are fit to the data?
b. In Step 1, the variable $x_1$ is selected as the "best" one-variable predictor. How is this determined?
c. In Step 2 of the stepwise regression, how many different two-variable models (where $x_1$ is one of the variables) are fit to the data?
d. The only two variables selected for entry into the stepwise regression model were $x_1$ and $x_8$. The stepwise regression yielded the following prediction equation:
$$
\hat y = .12 - .28x_1 + .27x_8.
$$
Give a practical interpretation of the $\beta$ estimates multiplied by $x_1$ and $x_8$.
e. Why should a researcher be wary of using the model, part (d), as the final model for predicting effort ($y$)?
:::


::: {.answer}
a. There are eight diﬀerent one-variable models that can be ﬁt to the data.
b. t-statistic is larger (absolute) than any of the other t-values. Or equivalently, p-value is smaller than any of the other p-values.
c. There would be an additional 7 two-variable combinations that could be ﬁtted to the model with $x_1$ being one of the variables.
d.

- $\hat\beta_1 = −0.28$. The difference in the mean relative error in estimating effort ($y$) between $x_1=0$ and $1$ is estimated to be $-0.28$. 
- $\hat\beta_2 = 0.27$. The difference in the mean relative error in estimating effort ($y$) between $x_8=0$ and $1$ is estimated to be $0.27$.
e. One should be wary of using the model in part d as the ﬁnal model because it contains only indicator variables and only ﬁrst-order terms.
:::


::: {#exr-}
## Dataset `C.csv`

The dataset `C.csv` contains 7 independent variables and 1 response variable $Y$.

a. Conduct a stepwise regression analysis of the data using an available statistical software package.
b. Interpret the $\beta$ estimates in the resulting stepwise model.
c. What are the dangers associated with drawing inferences from the stepwise model?

::: 


::: {.answer}

```{r}
dat <- read.csv("C.csv")
fit <- lm(Y~., data=dat)
library(olsrr)
ols_step_both_p(fit, pent=0.05, prem=0.05)
```


c. This model was selected as the best model after many diﬀerent models were ﬁt to the data. There are two main dangers associated with using and interpreting this model:

- The overall probability with making at least one Type I error is extremely high.
- This model only includes linear components. Before using this model, the analyst should consider both quadratic and interaction relationship in the model.
:::



::: {#exr-}
## Dataset `G.csv`
Consider the dataset `G.csv`. The response variable is $y$.

a. Use stepwise regression (with stepwise selection) to find the best predictors of $y$.
b. Use stepwise regression (with backward elimination) to find the best predictors of $y$.
c. Use all-possible-regressions-selection to find the best predictors of $y$.
d. Compare the results, parts a-c. Which independent variables consistently are selected as the best predictors?
e. Explain how you would use the results, parts a-c, to develop a model for $y$.
:::


::: {.answer}

```{r}
dat <- read.csv("G.csv")
fit <- lm(y ~ ., data = dat)
```

```{r}
stepwise.turbine <- ols_step_both_p(fit, pent = 0.05, prem = 0.05)
stepwise.turbine$model
```


```{r}
backward.turbine <- ols_step_backward_p(fit, prem = 0.05)
backward.turbine$model
```


```{r}
library(leaps)
bestsubset <- regsubsets(y ~ ., data = dat, nvmax = 9)
bestsubset.summary <- summary(bestsubset)
bestsubset.summary$outmat
```

```{r}
#| fig-show: hold
rsq <- bestsubset.summary$rsq
adjrsq <- bestsubset.summary$adjr2
cp <- bestsubset.summary$cp
rss <- bestsubset.summary$rss
p <- seq(1, 9, 1)
par(mfrow = c(2, 2))
plot(p, rsq)
plot(p, adjrsq)
plot(p, cp)
plot(p, rss)
par(mfrow = c(1, 1))
```


d. Based on the results from a-c above, the four variables ($x_1$, $x_3$, $x_4$, and $x_5$) are constantly selected.
e. We would use $x_1$, $x_3$, $x_4$, and $x_5$ as independent variables for predicting $y$ in a multiple regression model, then checking to see if including interaction and higher-order terms would improve the model.
:::