
# Basics 






## Population vs Sample

- **Population**: the set of all elements of interest in a particular study.
- **Sample**: is a subset of the population
- **Statistical Inference**: The process of using data collected on a sample to draw conclusions about a population is called statistical inference. 




::: {.callout-tip collapse="true"}
## Example: `gssr`
According to a previous national media report, the *average age* of U.S. adults who watch national TV news is **60 years**. A data analyst hypothesizes that the **average age of frequent TV-news watchers** is **greater than 60**. To test this, she uses the **General Social Survey (GSS)**, accessible via the `gssr` R package @gssr, to draw a **sample of 500 respondents** and record their ages.

a. **Describe the population**: All U.S. adults who watch TV news **several times a week or more often**, as defined by the GSS survey question on TV-news frequency.
b. **Describe the sample**: A random sample of 500 GSS respondents who meet the above criterion and have valid age data.
c. **Sample size**: $n = 500$
d. **Describe the statistical inference**: We want to determine whether the mean age of frequent TV-news watchers is greater than 60 based on the sampled data.
:::



## Data structure
A dataset is a collection of *values*. Values are organized in two ways. Every value belongs to a *variable* and an *observation*. 


::: {#def-}
# Variables and Observations @Wickham2014
- A **variable** contains all values that measure the same underlying attribute (like height, temperature, duration) across units.
- An **observation** contains all values measured on the same unit (like a person, or a day, or a race) across attributes.
:::
 


::: {.callout-tip collapse="true"}
## Example: `possum`
Researchers collected body measurements for bushtail possums in Eastern Australia. They trapped 104 possums and recorded location where possum was trapped, age, gender, head length, skull width, total length, tail length, etc. for each possum. For more details see @sec-app_possum.

```{r}
#| warning: false
library(DAAG)
data(possum)
head(possum)
```
The dimension of the dataset is
```{r}
dim(possum)
```

In this example, there are 14 varaibles, and 104 observations. Each column is a variable. In R you can use the column name to get access to the variable.

```{r}
possum$hdlngth
```

:::





::: {#def-}
# Quantitative and Qualitative @Men2020
- **Quantitative data** are observations measured on a naturally occurring numerical scale. It is also called *numerical data*.
- **Qualitative data** are nonnumerical data that can only be classified into one of a group of categories. It is also called *categorical data*.
:::



::: {#def-}
# Discrete and Continuous
- Discrete random variable: A random variable that assumes either a finite number of values or an infinite sequence of values such as 0, 1, 2...
- Continuous random variable: A random variable that may assume any numerical value in an interval or collection of intervals.
:::

```{mermaid}
graph TD
    A[All Variables] --> B[Numerical]
    A --> C[Categorical]
    B --> D[Continuous]
    B --> E[Discrete]
```


::: {.callout-note}
# Random variables
A variable in a dataset can be modeled by a *random variable*. The probability density function / probability mass function of the random variable can describe the distribution of all possible values of the variable in a dataset. 

Then to make a mearuement is the same as to take a sample from the random variable.  

:::






## Data Visualization
### Qualitative (categorical) data
Usually the most important is the **class relative frequency**: 
$$
\text{class relative frequency}=\frac{\text{class frequency}}{n}.
$$

To display it, we could use table, bar chart or pie chart.


::: {#exm- }




## Example: `possum`
Here we still consider the `possum` dataset. 

```{r}
library(DAAG)
data(possum)
head(possum)
```
The variable `Pop` (location where possum was trapped) is qualitative. We could use the following ways to display it. We first compute the frequency table of the variable `Pop`:

```{r}
table(possum$Pop)
```
or relative frequency talbe:
```{r}
table(possum$Pop)/length(possum$Pop)
```

Then we could draw the barplot of this variable:

```{r}
barplot(table(possum$Pop))
```

and the pie plot:
```{r}
pie(table(possum$Pop))
```

Note that `table` is handling the statistics, while `barplot` and `pie` draw on top of the result from `table`.

:::

### Quantitative (numerical) data


For a quantitative data, we would like to see the histogram, as well as computing some statistics: minx, max, quartiles, median and mean.

- **Histogram** is an approximation of the distribution. In other words, we split the range into small segments (called *bins*), and count the frequency or relative frequency of data falling into these bins.
- **Box plot** consists of a box, two lines and possibly some points:
    - The box in the box plot extends from the lower quartile to the upper quartile. The difference between the upper quartile and the lower quartile is called the inter-quartile range (IQR).
    - The lines, known as *whiskers*, extends to one and a half times the interquartile range, but then they are limited to reaching actual data points. 
    - The points, considered as *outliers*, are those which are not covered by the box and the lines.


::: {#exm-}
## `possum`
<details>
<summary><strong>Expand for details</strong></summary>

We still consider the `possum` dataset. We could use `summary` to compute the major statistics.

```{r}
library(DAAG)
data(possum)
summary(possum)
```

The variable `hdlngth` (head length) is quantitative. We first show the histogram. We could use `breaks` to control the number of bins. Note that the function `hist` does not just draw the histogram. It also provide many infomation we might need. 

```{r}
res <- hist(possum$hdlngth, breaks=10)
res
```

Then we could show the box plot. You may compare the lines in the plot with the summary table and the histogram above.

```{r}
boxplot(possum$hdlngth)
```


</details>
:::


### Relations among multiple variables

We could show the relation between two variables in a scatter plot.


::: {.callout-tip collapse="true"}
## Example: `possum`

The cases that both variables are numerical continuous:
```{r}
plot(possum$hdlngth, possum$skullw)
```

The cases that one variable is categorical:

```{r}
plot(as.factor(possum$Pop), possum$skullw)
```

Note that in this case, the categorical data has to be a factor. And once it is cast into a factor, the plot is multiple box plots for each category.


We can see pairwise plots for each pair of variables. Note that before the plot, we have to cast `pop` and `sex` into factors.

```{r}
possum$Pop <- as.factor(possum$Pop)
possum$sex <- as.factor(possum$sex)
pairs(possum)
```
Pair plot is very important since it can easily help us to find the expected relations between variables. We will use it a lot in regression analysis.

:::


## Probability Distribution

### Normal Probability Distribution

::: {#thm-norm}
## Theorem (Normal Sample Mean–Variance)
Let $X_1,\ldots,X_n\sim N(\mu,\sigma^2)$ i.i.d. Consider

- $\bar X=\frac1n\sum_{i=1}^nX_i$
- $s^2=\frac1{n-1}\sum_{i=1}^n(X_i-\bar X)^2$

Then

- $\bar X\sim N(\mu,\frac{\sigma^2}n)$
- $\frac{(n-1)s^2}{\sigma^2}\sim \chi^2_{n-1}$
- $\bar X$ and $s^2$ are independent.
:::



### Student's t-Distribution


@thm-t

@thm-norm

::: {#thm-t}
## Theorem (Student's t-Distribution)
Let 

- $Z\sim N(0,1)$ be a standard normal random variable
- $U\sim \chi_{\nu}^2$ be a chi-square random variable with $\nu$ degrees of freedom. 
- $Z$ and $U$ are independent.

Then
$$
T=\frac{Z}{\sqrt{U/\nu}}
$$
has a Student's t-distribution with $\nu$ degrees of freedom: $T\sim t_{\nu}$.
:::