
# Basics 






## Population vs Sample

- **Population**: the set of all elements of interest in a particular study.
- **Sample**: is a subset of the population
- **Statistical Inference**: The process of using data collected on a sample to draw conclusions about a population is called statistical inference. 




::: {.callout-tip collapse="true"}
## Example: `gssr`
According to a previous national media report, the *average age* of U.S. adults who watch national TV news is **60 years**. A data analyst hypothesizes that the **average age of frequent TV-news watchers** is **greater than 60**. To test this, she uses the **General Social Survey (GSS)**, accessible via the `gssr` R package @gssr, to draw a **sample of 500 respondents** and record their ages.

a. **Describe the population**: All U.S. adults who watch TV news **several times a week or more often**, as defined by the GSS survey question on TV-news frequency.
b. **Describe the sample**: A random sample of 500 GSS respondents who meet the above criterion and have valid age data.
c. **Sample size**: $n = 500$
d. **Describe the statistical inference**: We want to determine whether the mean age of frequent TV-news watchers is greater than 60 based on the sampled data.
:::



## Data structure
A dataset is a collection of *values*. Values are organized in two ways. Every value belongs to a *variable* and an *observation*. 


::: {#def-}
# Variables and Observations @Wickham2014
- A **variable** contains all values that measure the same underlying attribute (like height, temperature, duration) across units.
- An **observation** contains all values measured on the same unit (like a person, or a day, or a race) across attributes.
:::
 


::: {.callout-tip collapse="true"}
## Example: `possum`
Researchers collected body measurements for bushtail possums in Eastern Australia. They trapped 104 possums and recorded location where possum was trapped, age, gender, head length, skull width, total length, tail length, etc. for each possum. For more details see @sec-app_possum.

```{r}
#| warning: false
library(DAAG)
data(possum)
head(possum)
```
The dimension of the dataset is
```{r}
dim(possum)
```

In this example, there are 14 varaibles, and 104 observations. Each column is a variable. In R you can use the column name to get access to the variable.

```{r}
possum$hdlngth
```

:::





::: {#def-}
# Quantitative and Qualitative @Mendenhall2020
- **Quantitative data** are observations measured on a naturally occurring numerical scale. It is also called *numerical data*.
- **Qualitative data** are nonnumerical data that can only be classified into one of a group of categories. It is also called *categorical data*.
:::



::: {#def-}
# Discrete and Continuous
- Discrete random variable: A random variable that assumes either a finite number of values or an infinite sequence of values such as 0, 1, 2...
- Continuous random variable: A random variable that may assume any numerical value in an interval or collection of intervals.
:::

```{mermaid}
graph TD
    A[All Variables] --> B[Numerical]
    A --> C[Categorical]
    B --> D[Continuous]
    B --> E[Discrete]
```


::: {.callout-note}
# Random variables
A variable in a dataset can be modeled by a *random variable*. The probability density function / probability mass function of the random variable can describe the distribution of all possible values of the variable in a dataset. 

Then to make a mearuement is the same as to take a sample from the random variable.  

:::






## Data Visualization
### Qualitative (categorical) data
Usually the most important is the **class relative frequency**: 
$$
\text{class relative frequency}=\frac{\text{class frequency}}{n}.
$$

To display it, we could use table, bar chart or pie chart.


::: {.callout-tip collapse="true"}




## Example: `possum`

All the example about possums below come from the dataset [here](./assests/data/possum.csv).


::: {.panel-tabset}
## R
```{r}
df <- read.csv('assests/data/possum.csv')
head(df)
```

We first compute the frequency table of the variable `pop`:

```{r}
table(df$pop)
```
or relative frequency talbe:
```{r}
table(df$pop)/length(df$pop)
```

Then we could draw the barplot of this variable:

```{r}
barplot(table(df$pop))
```

and the pie plot:
```{r}
pie(table(df$pop))
```

Note that `table` is handling the statistics, while `barplot` and `pie` draw on top of the result from `table`.


:::
:::

### Quantitative (numerical) data

We would like to use **histogram** to display these type of data. In other words, we split the range into small segments (called *bins*), and count the frequency or relative frequency of data falling into these bins.


::: {.callout-tip collapse="true"}
## Example: `possum`
```{r}
df <- read.csv('assests/data/possum.csv')
head(df)
```

We display the histogram of `headL`. We could use `breaks` to control the number of bins.
```{r}
hist(df$headL, breaks=10)
```

The function `hist` does not just draw the histogram. It also provide many infomation we might need. 

```{r}
res <- hist(df$headL, breaks=10)
res
```

:::

### Some statistics

For a quantitative data, we would also like to compute some statistics: minx, max, quartiles, median and mean. In R, we could use `summary` to compute them, and use box plots to show them.


::: {.callout-tip collapse="true"}
## Example: `possum`


```{r}
summary(df)
```

We could use the box plot to show these infomation. It consists of a box, two lines and possibly some points:

- The box in the box plot extends from the lower quartile to the upper quartile. The difference between the upper quartile and the lower quartile is called the inter-quartile range (IQR).
- The lines, known as *whiskers*, extends to one and a half times the interquartile range, but then they are limited to reaching actual data points. 
- The points, considered as *outliers*, are those which are not covered by the box and the lines.


```{r}
boxplot(df$headL)
```


:::

### Relations among multiple variables

We could show the relation between two variables in a scatter plot.


::: {.callout-tip collapse="true"}
## Example: `possum`

The cases that both variables are numerical continuous:
```{r}
plot(df$headL, df$skullW)
```

The cases that one variable is categorical:

```{r}
plot(as.factor(df$pop), df$skullW)
```

Note that in this case, the categorical data has to be a factor. And once it is cast into a factor, the plot is multiple box plots for each category.


We can see pairwise plots for each pair of variables. Note that before the plot, we have to cast `pop` and `sex` into factors.

```{r}
df$pop <- as.factor(df$pop)
df$sex <- as.factor(df$sex)
pairs(df)
```
Pair plot is very important since it can easily help us to find the expected relations between variables. We will use it a lot in regression analysis.

:::



## p


::: {#def-}
# Population
A **population data set** is a collection (or set) of data measured on all experimental units of interest to you. @Men2020
:::

::: {#def-}
# Sample
A **sample** is a subset of data selected from a population. @Men2020
:::



::: {#def-}
# random sample
A **random sample** of $n$ experimental units is one selected from the population in such a way that every different sample of size $n$ has an equal probability of selection. @Men2020
:::





### Inferential statistics


::: {#def-}
# Statistical inference @Men2020
- A **statistical inference** is an estimate, prediction, or some other generatlization about a population based on information contianed in a sample. 
- A **measure of reliability** is a statement about the degree of uncertainty associated with a statistical inference.
:::



::: {.callout-note}
# Inferential statistics
1. Identify **population**
2. Identify **variable(s)**
3. Collect **sample** data
4. **Inference** about population based on sample
5. **Measure** of reliability for inference
:::


