


























## Example: One-Way ANOVA  

Suppose a researcher wants to compare the effectiveness of three teaching methods on exam performance. A group of students is randomly assigned to one of three methods, and their exam scores are recorded.

| Method | Scores |
|--------|--------|
| A | 78, 82, 85, 80 |
| B | 88, 90, 85, 87 |
| C | 72, 75, 70, 74 |

We want to test whether the mean scores differ across methods:

$$
H_0:\ \mu_A=\mu_B=\mu_C
\qquad\text{vs}\qquad
H_a:\ \text{not all means are equal}.
$$

First compute the sample means:

$$
\bar y_A=81.25,\quad 
\bar y_B=87.5,\quad 
\bar y_C=72.75.
$$

We observe differences among the group means, but ANOVA determines whether these differences are large relative to the variation within each group.

The one-way ANOVA model is

$$
y_{ij}=\mu+\tau_i+\varepsilon_{ij},\qquad i=1,2,3,
$$

where $\tau_i$ represents the effect of teaching method $i$.

ANOVA partitions variability into:

- **Between-group variation**: differences among group means  
- **Within-group variation**: variability of observations within each group  

If the between-group variation is sufficiently large relative to the within-group variation, we reject $H_0$ and conclude that at least one teaching method differs from the others.



```{r}
# Example data
score <- c(
    78, 82, 85, 80,
    88, 90, 85, 87,
    72, 75, 70, 74
)

method <- factor(c(
    "A", "A", "A", "A",
    "B", "B", "B", "B",
    "C", "C", "C", "C"
))

dat <- data.frame(method, score)

# View the data
dat

# Group means
aggregate(score ~ method, data = dat, mean)

# Fit one-way ANOVA
fit <- aov(score ~ method, data = dat)

# ANOVA table
summary(fit)

# Equivalent regression model
lm_fit <- lm(score ~ method, data = dat)
summary(lm_fit)
anova(lm_fit)

# Optional: boxplot
boxplot(score ~ method,
    data = dat,
    main = "Exam Scores by Teaching Method",
    xlab = "Method",
    ylab = "Score"
)
```




















::: {#exm-}
Suppose we study exam scores from three teaching methods.

```{r}
set.seed(123)

method <- factor(rep(c("Lecture", "Hybrid", "Online"), each = 12))
score <- c(
  rnorm(12, mean = 78, sd = 4),
  rnorm(12, mean = 84, sd = 4),
  rnorm(12, mean = 80, sd = 4)
)

fit <- lm(score ~ method)
summary(fit)
model.tables(aov(score ~ method), type = "means")
```

If `Lecture` is the reference level, then:

- the intercept estimates the mean score for `Lecture`;
- the coefficient of `methodHybrid` estimates the difference between `Hybrid` and `Lecture`;
- the coefficient of `methodOnline` estimates the difference between `Online` and `Lecture`.

The overall F-test checks whether all group means are equal.
:::


## How R handles categorical variables

In R, factors are used to represent categorical predictors. When a factor is included in `lm()`, R automatically creates the corresponding dummy variables.

```{r}
#| eval: false

dat$group <- as.factor(dat$group)
contrasts(dat$group)
fit <- lm(y ~ group, data = dat)
summary(fit)
```

If we want to change the baseline group, we can use `relevel()`.

```{r}
#| eval: false

dat$group <- relevel(dat$group, ref = "control")
fit <- lm(y ~ group, data = dat)
summary(fit)
```


::: {.callout-tip}
Changing the reference level does not change the fitted values, residuals, SSE, or the overall F-test. It only changes how the coefficients are written and interpreted.
:::







## Relation to one-way ANOVA

The traditional one-way ANOVA model is

$$
y_{ij}=\mu+\tau_i+\varepsilon_{ij},
$$

where $\tau_i$ is the effect of group $i$.

This is equivalent to a regression model with dummy variables. For example, with three groups:

$$
y=\beta_0+\beta_1x_1+\beta_2x_2+\varepsilon.
$$

These two formulations describe the same fitted means, the same residuals, and the same F-test for

$$
H_0:\mu_1=\mu_2=\cdots=\mu_g.
$$

In regression language, this is the same as testing

$$
H_0:\beta_1=\beta_2=\cdots=\beta_{g-1}=0.
$$

Therefore, one-way ANOVA is exactly categorical regression with one factor.


::: {.callout-note}
The classical ANOVA table partitions variability into between-group and within-group components. In the regression formulation, this is just the usual decomposition into explained and unexplained variation.
:::



## Historical note: ANOVA and regression

Historically, ANOVA and regression were developed as closely related but somewhat separate traditions.

- Regression was developed to study relationships among quantitative variables.
- ANOVA was developed to compare means across groups, especially in designed experiments.

In modern linear model theory, these are unified. Both are special cases of the same model

$$
\mathbf y=\mathbf X\boldsymbol{\beta}+\boldsymbol{\varepsilon}.
$$

The difference is mainly in how the columns of the design matrix $\mathbf X$ are constructed:

- quantitative regression uses numerical predictors directly;
- categorical regression uses indicator variables;
- ANOVA tables summarize hypothesis tests for groups of coefficients by comparing nested models.

So traditional ANOVA is not really a separate method from regression. It is a special presentation of linear-model inference when predictors are categorical.

