


























## Example: One-Way ANOVA  

Suppose a researcher wants to compare the effectiveness of three teaching methods on exam performance. A group of students is randomly assigned to one of three methods, and their exam scores are recorded.

| Method | Scores |
|--------|--------|
| A | 78, 82, 85, 80 |
| B | 88, 90, 85, 87 |
| C | 72, 75, 70, 74 |

We want to test whether the mean scores differ across methods:

$$
H_0:\ \mu_A=\mu_B=\mu_C
\qquad\text{vs}\qquad
H_a:\ \text{not all means are equal}.
$$

First compute the sample means:

$$
\bar y_A=81.25,\quad 
\bar y_B=87.5,\quad 
\bar y_C=72.75.
$$

We observe differences among the group means, but ANOVA determines whether these differences are large relative to the variation within each group.

The one-way ANOVA model is

$$
y_{ij}=\mu+\tau_i+\varepsilon_{ij},\qquad i=1,2,3,
$$

where $\tau_i$ represents the effect of teaching method $i$.

ANOVA partitions variability into:

- **Between-group variation**: differences among group means  
- **Within-group variation**: variability of observations within each group  

If the between-group variation is sufficiently large relative to the within-group variation, we reject $H_0$ and conclude that at least one teaching method differs from the others.



```{r}
# Example data
score <- c(
    78, 82, 85, 80,
    88, 90, 85, 87,
    72, 75, 70, 74
)

method <- factor(c(
    "A", "A", "A", "A",
    "B", "B", "B", "B",
    "C", "C", "C", "C"
))

dat <- data.frame(method, score)

# View the data
dat

# Group means
aggregate(score ~ method, data = dat, mean)

# Fit one-way ANOVA
fit <- aov(score ~ method, data = dat)

# ANOVA table
summary(fit)

# Equivalent regression model
lm_fit <- lm(score ~ method, data = dat)
summary(lm_fit)
anova(lm_fit)

# Optional: boxplot
boxplot(score ~ method,
    data = dat,
    main = "Exam Scores by Teaching Method",
    xlab = "Method",
    ylab = "Score"
)
```




















::: {#exm-}
Suppose we study exam scores from three teaching methods.

```{r}
set.seed(123)

method <- factor(rep(c("Lecture", "Hybrid", "Online"), each = 12))
score <- c(
  rnorm(12, mean = 78, sd = 4),
  rnorm(12, mean = 84, sd = 4),
  rnorm(12, mean = 80, sd = 4)
)

fit <- lm(score ~ method)
summary(fit)
model.tables(aov(score ~ method), type = "means")
```

If `Lecture` is the reference level, then:

- the intercept estimates the mean score for `Lecture`;
- the coefficient of `methodHybrid` estimates the difference between `Hybrid` and `Lecture`;
- the coefficient of `methodOnline` estimates the difference between `Online` and `Lecture`.

The overall F-test checks whether all group means are equal.
:::


## How R handles categorical variables

In R, factors are used to represent categorical predictors. When a factor is included in `lm()`, R automatically creates the corresponding dummy variables.

```{r}
#| eval: false

dat$group <- as.factor(dat$group)
contrasts(dat$group)
fit <- lm(y ~ group, data = dat)
summary(fit)
```

If we want to change the baseline group, we can use `relevel()`.

```{r}
#| eval: false

dat$group <- relevel(dat$group, ref = "control")
fit <- lm(y ~ group, data = dat)
summary(fit)
```


::: {.callout-tip}
Changing the reference level does not change the fitted values, residuals, SSE, or the overall F-test. It only changes how the coefficients are written and interpreted.
:::







## Relation to one-way ANOVA

The traditional one-way ANOVA model is

$$
y_{ij}=\mu+\tau_i+\varepsilon_{ij},
$$

where $\tau_i$ is the effect of group $i$.

This is equivalent to a regression model with dummy variables. For example, with three groups:

$$
y=\beta_0+\beta_1x_1+\beta_2x_2+\varepsilon.
$$

These two formulations describe the same fitted means, the same residuals, and the same F-test for

$$
H_0:\mu_1=\mu_2=\cdots=\mu_g.
$$

In regression language, this is the same as testing

$$
H_0:\beta_1=\beta_2=\cdots=\beta_{g-1}=0.
$$

Therefore, one-way ANOVA is exactly categorical regression with one factor.


::: {.callout-note}
The classical ANOVA table partitions variability into between-group and within-group components. In the regression formulation, this is just the usual decomposition into explained and unexplained variation.
:::



## Historical note: ANOVA and regression

Historically, ANOVA and regression were developed as closely related but somewhat separate traditions.

- Regression was developed to study relationships among quantitative variables.
- ANOVA was developed to compare means across groups, especially in designed experiments.

In modern linear model theory, these are unified. Both are special cases of the same model

$$
\mathbf y=\mathbf X\boldsymbol{\beta}+\boldsymbol{\varepsilon}.
$$

The difference is mainly in how the columns of the design matrix $\mathbf X$ are constructed:

- quantitative regression uses numerical predictors directly;
- categorical regression uses indicator variables;
- ANOVA tables summarize hypothesis tests for groups of coefficients by comparing nested models.

So traditional ANOVA is not really a separate method from regression. It is a special presentation of linear-model inference when predictors are categorical.








## Relation to two-way ANOVA

Suppose there are two categorical predictors, say factor A and factor B. Then the two-way ANOVA model without interaction is

$$
y_{ijk}=\mu+\alpha_i+\beta_j+\varepsilon_{ijk}.
$$

This corresponds to a regression model containing dummy variables for factor A and dummy variables for factor B.

If we include interaction, the model becomes

$$
y_{ijk}=\mu+\alpha_i+\beta_j+(\alpha\beta)_{ij}+\varepsilon_{ijk}.
$$

In regression form, this means:

1. include dummy variables for factor A;
2. include dummy variables for factor B;
3. include products of those dummy variables to represent the interaction.

Thus:

- two-way ANOVA without interaction corresponds to an additive regression model;
- two-way ANOVA with interaction corresponds to a regression model with interaction terms between categorical predictors.


::: {#exm-}
Here is a small two-way example.

```{r}
set.seed(321)

A <- factor(rep(c("Low", "High"), each = 18))
B <- factor(rep(rep(c("Morning", "Evening", "Night"), each = 6), times = 2))

y <- c(
  rnorm(6, 70, 3), rnorm(6, 74, 3), rnorm(6, 72, 3),
  rnorm(6, 78, 3), rnorm(6, 85, 3), rnorm(6, 82, 3)
)

fit.add <- lm(y ~ A + B)
fit.int <- lm(y ~ A * B)

anova(fit.add, fit.int)
```

The nested-model comparison checks whether the interaction terms are needed. If the interaction is not significant, the additive model may be adequate. If the interaction is significant, the effect of one factor depends on the level of the other factor.
:::


## Traditional ANOVA language versus regression language

The two vocabularies are slightly different, but they refer to the same underlying model.

| Traditional ANOVA language | Regression language |
|---|---|
| treatment or factor | categorical predictor |
| level | category |
| main effect | coefficients associated with one factor |
| interaction | product terms |
| compare means | test linear combinations of coefficients |
| between-group variation | explained variation due to factor terms |
| within-group variation | residual variation |


## Summary

- Categorical regression is ordinary linear regression with dummy variables.
- A one-way ANOVA model is a regression model with one categorical predictor.
- A two-way ANOVA model is a regression model with two categorical predictors, possibly with interaction.
- The ANOVA F-tests are nested-model tests on groups of regression coefficients.
- Modern linear model theory unifies regression and ANOVA in one framework.
