# Example: Boston Housing dataset full walkthrough

In the previous section we talk about our main tools to handle the `Boston` dataset. In this section we display how to explore the dataset from scratch.



| Variable  | Description                                                                 | Type                    |
|-----------|-----------------------------------------------------------------------------|-------------------------|
| `crim`    | Per capita crime rate by town                                                | Numerical, continuous   |
| `zn`      | Proportion of residential land zoned for lots over 25,000 sq.ft.            | Numerical, continuous   |
| `indus`   | Proportion of non-retail business acres per town                             | Numerical, continuous   |
| `chas`    | Charles River dummy variable (1 if tract bounds river; 0 otherwise)          | Categorical, nominal   |
| `nox`     | Nitrogen oxides concentration (parts per 10 million)                        | Numerical, continuous   |
| `rm`      | Average number of rooms per dwelling                                         | Numerical, continuous   |
| `age`     | Proportion of owner-occupied units built prior to 1940                       | Numerical, continuous   |
| `dis`     | Weighted mean distance to five Boston employment centers                     | Numerical, continuous   |
| `rad`     | Index of accessibility to radial highways (larger = better accessibility)   | Categorical, ordinal   |
| `tax`     | Full-value property-tax rate per \$10,000                                    | Numerical, continuous   |
| `ptratio` | Pupil–teacher ratio by town                                                  | Numerical, continuous   |
| `black`   | $1000(B_k - 0.63)^2$, where $B_k$ is the proportion of Black residents   | Numerical, continuous   |
| `lstat`   | Percentage of lower-status population                                       | Numerical, continuous   |
| `medv`    | Median value of owner-occupied homes (in \$1,000s)                          | Numerical, continuous   |



## Phase 1: Understand the Data Before Fitting Anything
Before touching `lm()`, the first step is always exploratory data analysis (EDA). 


```{r}
library(MASS)
library(car)
data(Boston)

Boston$chas <- factor(Boston$chas)
Boston$rad <- factor(Boston$rad)

summary(Boston)
```


```{r}
pairs(Boston)
```

Correlation among numeric variables

```{r}
#| warning: false
num_vars <- Boston[, sapply(Boston, is.numeric)]
cor_matrix <- cor(num_vars)

library(ggcorrplot)
ggcorrplot(cor_matrix, hc.order = TRUE)
```

What I'm looking for:

- Skewed variables (may need log transforms)
- Non-linear relationships between predictors and `medv`
- High pairwise correlations between predictors (multicollinearity warning signs)
- Outliers or influential observations

Here are some key takeaways from the observations:

- `lstat` vs `medv`: curved, which indicates quadratic or log transform likely needed
- `crim` is extremely right-skewed, which indicates `log(crim)` often works better
- `rm` vs `medv`: mostly linear but with some spread
- `nox`, `dis`, `age` are highly intercorrelated (all related to urban density)


## Phase 2: Start with All Predictors — Full Model

We begin with the full additive regression model.

```{r}
fit_full <- lm(medv ~ ., data = Boston)
summary(fit_full)
```
This model provides a baseline for comparison. 

```{r}
#| fig-show: hold
par(mfrow = c(2, 2))
plot(fit_full)
par(mfrow = c(1, 1))
```

Here are some important info we get. 

- $R_{adj}^2\approx0.74$ and $R^2\approx0.75$. So about $74-75\%$ of the variability in median house value is explained by the predictors.
- Overall p-value for the overall F-test is very small, which means that the whole group of variables work.
- `indus`, `age` and some of the `rad` levels are not signicant.

Then immediately do residual diagnostics on this full model:


The four diagnostic plots give me:

1. Residuals vs Fitted — checks linearity and homoscedasticity
2. Normal Q-Q — checks normality of residuals
3. Scale-Location — another view of heteroscedasticity
4. Residuals vs Leverage — identifies influential points (Cook's distance)

On raw Boston data, these plots typically show a fan shape (heteroscedasticity) and non-linearity — which tells me transformations are needed before I proceed.

### Variable selections

We apply best-of-subset to select variables.


```{r}
library(leaps)
fits <- regsubsets(
    medv ~ .,
    data = Boston,
    nvmax = ncol(model.matrix(medv ~ ., Boston)) - 1
)
models_summary <- summary(fits)
best_cp <- which.min(models_summary$cp)
plot(models_summary$cp, xlab = "Number of varaiables", type = "l")
points(best_cp, models_summary$cp[best_cp], col = "red", cex = 2, pch = 20)

models_summary
```

Since for dummy variable `rad` we have to choose all or none, based on the best-of-subset results, we will exclude `indus` and `age` from our model. We will delete the two variables from `Boston` and name the dataset `dat`.


```{r}
dat <- subset(Boston, select = -c(indus, age))
```


Therefore our based model is 
```{r}
fit_base <- lm(medv~., data=dat)
summary(fit_base)
```

The corresponding residual plots are
```{r}
par(mfrow = c(2, 2))
plot(fit_base)
par(mfrow = c(1, 1))
```




## Phase 3: Consider Transformations





### Check if `log(medv)` improves things

Based on all plots, we may need to deal with several curvy patterns as well as the non-constant variance issue. Therefore we would like to start from a transformation for `medv`. I'd also use Box-Cox to formally choose the best power transformation for `medv`:

```{r}
boxcox(fit_base)  
```

Reading this specific plot

- The peak appears to be right around $\lambda\approx0$, and 
- the 95% CI dotted lines straddle 0 closely. They look to be roughly $(-0.1, 0.2)$ or so.

By the standard rule, if $\lambda=0$ is inside the 95% CI, `log(y)` is the transformation.

```{r}
fit_log <- lm(log(medv) ~ ., data = dat)
par(mfrow = c(2, 2))
plot(fit_log)
par(mfrow = c(1, 1))
```

```{r}
summary(fit_log)
```

With `log(medv)` as the response:

- Coefficients are now interpreted as percent changes
- The residual fan shape (heteroscedasticity) is slightly improved


### Partial residual plots

Then we look at the partial residual plots.

First show the partial residual plots.

```{r}
par(mfrow = c(4, 3))
crPlots(fit_log)
par(mfrow = c(1, 1))
```

Based on the plots, it seems that `crim`, `rm`, `dis` and `lstat` is a little curvy. Therefore we would like to try squared terms for them.


```{r}
fit2 <- lm(log(medv)~.+I(crim^2)+I(rm^2)+I(dis^2)+I(lstat^2), data=dat)
summary(fit2)
par(mfrow = c(4, 3))
crPlots(fit2)
par(mfrow = c(1, 1))
```

The quadratic terms improve the model substantially. Compared with `fit_log`, the residual standard error drops from about `0.189` to about `0.171`, and the nested F-test for adding the four squared terms is highly significant.

```{r}
anova(fit_log, fit2)
```

At this point, `zn` is no longer significant after the nonlinear terms are added, so we can simplify the model by dropping it and refitting.

```{r}
fit3 <- update(fit2, . ~ . - zn)
summary(fit3)
```

This simpler model has essentially the same fit as `fit2`, so removing `zn` costs us almost nothing while making the model easier to explain.

```{r}
anova(fit3, fit2)
```

## Phase 4: Diagnose the Final Model

Now we return to the standard diagnostics for the refined model.

```{r}
#| fig-show: hold
par(mfrow = c(2, 2))
plot(fit3)
par(mfrow = c(1, 1))
```

These plots are much better than the ones from the original untransformed model:

- The residuals versus fitted plot shows less curvature.
- The spread is more stable across fitted values.
- The Q-Q plot is closer to a straight line, although the tails still deviate somewhat.
- A few observations remain influential, which is common in this dataset.

We can also check multicollinearity and influence numerically.

```{r}
vif(fit3)
```

The polynomial terms create some moderate to large VIF values, especially for `rm` with `I(rm^2)` and `dis` with `I(dis^2)`. This is expected when a variable and its square are both in the model, so it is not automatically a reason to remove them.

```{r}
rs <- rstudent(fit3)
cd <- cooks.distance(fit3)
lev <- hatvalues(fit3)

c(
  max_abs_studentized = max(abs(rs)),
  num_abs_studentized_gt_3 = sum(abs(rs) > 3),
  num_cooks_gt_4_over_n = sum(cd > 4 / nrow(dat)),
  num_high_leverage = sum(lev > 2 * length(coef(fit3)) / nrow(dat))
)
```

This tells us that there are still some unusual observations, but the overall model is far more stable than the original additive fit on the raw response. Unless those data points are recording errors, I would usually keep them and report their presence rather than delete them automatically.

## Final Model

Our final working model is

$$
\log(\texttt{medv}) = \beta_0 + \beta_1\texttt{crim} + \beta_2\texttt{chas}
+ \beta_3\texttt{nox} + \beta_4\texttt{rm} + \beta_5\texttt{dis}
+ \beta_6\texttt{rad} + \beta_7\texttt{tax} + \beta_8\texttt{ptratio}
+ \beta_9\texttt{black} + \beta_{10}\texttt{lstat}
+ \beta_{11}\texttt{crim}^2 + \beta_{12}\texttt{rm}^2
+ \beta_{13}\texttt{dis}^2 + \beta_{14}\texttt{lstat}^2 + \varepsilon.
$$

In R formula form, this is:

```{r}
formula(fit3)
```

The main conclusions are:

- Transforming the response with `log(medv)` is justified by the Box-Cox plot and improves variance stability.
- Adding quadratic terms for `crim`, `rm`, `dis`, and `lstat` captures important curvature that the additive linear model misses.
- After those nonlinear terms are included, `zn` is no longer needed.
- The final model explains about `83%` of the variation in `log(medv)`, with adjusted $R^2$ around `0.825`.

## Takeaway

This example shows a typical regression workflow:

1. Start with EDA and a full model.
2. Use residual plots and partial residual plots to detect problems.
3. Transform the response when the variance and shape suggest it.
4. Add nonlinear terms where the data indicate curvature.
5. Recheck diagnostics before settling on the final model.

The important point is that model building is iterative. We do not fit one regression and stop; instead, we repeatedly use diagnostics to decide how to improve the model.
