# Example: DAAG::Possum dataset full walkthrough

In this report we study the `DAAG::possum` data and build a regression model for total length, `totlngth`.
The goal is not just to fit one model, but to show a typical workflow:

1. explore the data,
2. fit a reasonable additive model,
3. use diagnostics to look for problems,
4. add interactions when the plots suggest non-additive structure,
5. refit and interpret the final model.

## Variables used in this report

We will focus on the following variables.

| Variable | Description | Type |
|----------|-------------|------|
| `totlngth` | Total length of the possum | Numerical, continuous |
| `hdlngth` | Head length | Numerical, continuous |
| `taill` | Tail length | Numerical, continuous |
| `footlgth` | Foot length | Numerical, continuous |
| `Pop` | Population group (`Vic` or `other`) | Categorical |
| `sex` | Sex (`f` or `m`) | Categorical |

## Phase 1: Set up the data

We keep only the variables needed for this analysis and remove the one observation with a missing `footlgth`.

```{r}
library(DAAG)
library(car)

data(possum)

dat <- na.omit(possum[, c("Pop", "sex", "hdlngth", "taill", "footlgth", "totlngth")])

summary(dat)
```

There are `r nrow(dat)` complete observations in the analysis data set.

## Phase 2: Exploratory data analysis

Before fitting a model, it is useful to see the marginal relationships among the variables.

```{r}
pairs(dat)
```

For the numeric variables, the sample correlations are:

```{r}
round(cor(dat[, c("hdlngth", "taill", "footlgth", "totlngth")]), 3)
```

From these plots and correlations, a few patterns stand out:

- `totlngth` is strongly related to both `hdlngth` and `taill`.
- `footlgth` is positively associated with `totlngth`, but the relationship looks weaker than the first two.
- `hdlngth` and `footlgth` are themselves related, so we should keep an eye on redundancy.
- The scatterplots suggest that the effect of one body measurement may depend on another, which hints at possible interaction terms.

## Phase 3: Start with an additive model

A natural first model is an additive regression using the two categorical variables and the three body measurements:

```{r}
fit0 <- lm(totlngth ~ Pop + sex + hdlngth + taill + footlgth, data = dat)
summary(fit0)
```

This is already a fairly good model. In the additive fit:

- `hdlngth` and `taill` are strongly significant,
- `sex` is significant,
- `Pop` is borderline,
- `footlgth` is not significant as a main effect.

That last point does not necessarily mean `footlgth` is useless. It may be involved through interaction terms instead of a simple additive contribution.

### Diagnostics for the additive model

```{r}
#| fig-show: hold
par(mfrow = c(2, 2))
plot(fit0)
par(mfrow = c(1, 1))
```

The ordinary residual plots do not show a severe variance problem, but they are not very sensitive to hidden interaction structure. For that, component-plus-residual plots are often more informative.

```{r}
#| warning: false
par(mfrow = c(2, 2))
crPlots(fit0)
par(mfrow = c(1, 1))
```

The partial residual plots suggest that the effects of `taill` and `footlgth` may change with `hdlngth`. In other words, the mean structure may not be purely additive.

## Phase 4: Add interaction terms

Based on the plots and on the subject matter, we try the model

$$
\texttt{totlngth} \sim \texttt{Pop} + \texttt{sex} + \texttt{hdlngth} * \texttt{taill} + \texttt{hdlngth} * \texttt{footlgth}.
$$

```{r}
fit1 <- lm(totlngth ~ Pop + sex + hdlngth * taill + hdlngth * footlgth, data = dat)
summary(fit1)
```

Compared with the additive model, this interaction model fits better:

```{r}
c(
    rse_additive = summary(fit0)$sigma,
    adj_r2_additive = summary(fit0)$adj.r.squared,
    aic_additive = AIC(fit0),
    rse_interaction = summary(fit1)$sigma,
    adj_r2_interaction = summary(fit1)$adj.r.squared,
    aic_interaction = AIC(fit1)
)
```

The interaction model has a smaller residual standard error, larger adjusted $R^2$, and smaller AIC. So the extra complexity is justified here.

## Phase 5: Center the continuous predictors for interpretation

When a model contains interactions, the main-effect coefficients are easier to interpret if we center the continuous variables first.

```{r}
dat$c_hd <- scale(dat$hdlngth, center = TRUE, scale = FALSE)
dat$c_tail <- scale(dat$taill, center = TRUE, scale = FALSE)
dat$c_foot <- scale(dat$footlgth, center = TRUE, scale = FALSE)

fit_final <- lm(
    totlngth ~ Pop + sex + c_hd * c_tail + c_hd * c_foot,
    data = dat
)

summary(fit_final)
```

After centering:

- the intercept is the expected total length for a female possum from `Vic` with average head, tail, and foot length,
- the coefficient of `c_hd` is the effect of head length when the other two continuous variables are at their means,
- the interaction `c_hd:c_tail` shows how the effect of tail length changes as head length changes,
- the interaction `c_hd:c_foot` shows how the effect of foot length changes as head length changes.

The signs of the interaction terms are informative:

- `c_hd:c_tail` is negative, so the contribution of tail length becomes smaller as head length increases.
- `c_hd:c_foot` is positive, so the contribution of foot length becomes larger as head length increases.

## Phase 6: Diagnose the final model

```{r}
#| fig-show: hold
par(mfrow = c(2, 2))
plot(fit_final)
par(mfrow = c(1, 1))
```

These plots suggest that the final model is reasonably adequate:

- there is no strong funnel shape,
- the Q-Q plot is acceptable for a sample of this size,
- there is no obvious systematic pattern left in the residuals,
- no single observation appears overwhelmingly dominant.

We can also check variance stability and outliers numerically.

```{r}
ncvTest(fit_final)
outlierTest(fit_final)
```

There is no strong evidence of non-constant variance, and no observation is flagged as an extreme outlier after Bonferroni adjustment.

For completeness, we can also look at VIF values.

```{r}
vif(fit_final)
```

Because this model contains interactions, `vif()` warns about higher-order terms. The values are still moderate enough that multicollinearity does not look severe after centering.

## Final model

The final model is

$$
\texttt{totlngth}
= \beta_0 + \beta_1\texttt{Pop} + \beta_2\texttt{sex}
+ \beta_3 c\_hd + \beta_4 c\_tail + \beta_5 c\_foot
+ \beta_6(c\_hd \cdot c\_tail)
+ \beta_7(c\_hd \cdot c\_foot) + \varepsilon.
$$

In R formula form:

```{r}
formula(fit_final)
```

## Takeaway

This example illustrates an important modeling lesson. A predictor can look weak in a simple additive model and still become useful once interactions are introduced. In the possum data, `footlgth` is not important as a plain main effect, but it becomes relevant through its interaction with `hdlngth`.

So the final conclusion is:

- total length increases with head length and tail length,
- possums in the `other` population tend to be shorter than those in `Vic`,
- male possums tend to be slightly shorter after adjusting for the other variables,
- the effects of tail length and foot length depend on head length, so an interaction model is more appropriate than a purely additive one.
