# Example: Boston Housing dataset full walkthrough

In the previous section we talk about our main tools to handle the `Boston` dataset. In this section we display how to explore the dataset from scratch.



| Variable  | Description                                                                 | Type                    |
|-----------|-----------------------------------------------------------------------------|-------------------------|
| `crim`    | Per capita crime rate by town                                                | Numerical, continuous   |
| `zn`      | Proportion of residential land zoned for lots over 25,000 sq.ft.            | Numerical, continuous   |
| `indus`   | Proportion of non-retail business acres per town                             | Numerical, continuous   |
| `chas`    | Charles River dummy variable (1 if tract bounds river; 0 otherwise)          | Categorical, nominal   |
| `nox`     | Nitrogen oxides concentration (parts per 10 million)                        | Numerical, continuous   |
| `rm`      | Average number of rooms per dwelling                                         | Numerical, continuous   |
| `age`     | Proportion of owner-occupied units built prior to 1940                       | Numerical, continuous   |
| `dis`     | Weighted mean distance to five Boston employment centers                     | Numerical, continuous   |
| `rad`     | Index of accessibility to radial highways (larger = better accessibility)   | Categorical, ordinal   |
| `tax`     | Full-value property-tax rate per \$10,000                                    | Numerical, continuous   |
| `ptratio` | Pupil–teacher ratio by town                                                  | Numerical, continuous   |
| `black`   | $1000(B_k - 0.63)^2$, where $B_k$ is the proportion of Black residents   | Numerical, continuous   |
| `lstat`   | Percentage of lower-status population                                       | Numerical, continuous   |
| `medv`    | Median value of owner-occupied homes (in \$1,000s)                          | Numerical, continuous   |



## Phase 1: Understand the Data Before Fitting Anything
Before touching `lm()`, the first step is always exploratory data analysis (EDA). 


```{r}
library(MASS)
data(Boston)

Boston$chas <- factor(Boston$chas)
Boston$rad <- factor(Boston$rad)

summary(Boston)
```


```{r}
pairs(Boston)
```

Correlation among numeric variables

```{r}
#| warning: false
num_vars <- Boston[, sapply(Boston, is.numeric)]
cor_matrix <- cor(num_vars)

library(ggcorrplot)
ggcorrplot(cor_matrix, hc.order = TRUE)
```

What I'm looking for:

- Skewed variables (may need log transforms)
- Non-linear relationships between predictors and `medv`
- High pairwise correlations between predictors (multicollinearity warning signs)
- Outliers or influential observations

Here are some key takeaways from the observations:

- `lstat` vs `medv`: curved, which indicates quadratic or log transform likely needed
- `crim` is extremely right-skewed, which indicates `log(crim)` often works better
- `rm` vs `medv`: mostly linear but with some spread
- `nox`, `dis`, `age` are highly intercorrelated (all related to urban density)


## Phase 2: Start with All Predictors — Full Model

We begin with the full additive regression model.

```{r}
fit_full <- lm(medv ~ ., data = Boston)
summary(fit_full)
```
This model provides a baseline for comparison. 

```{r}
#| fig-show: hold
par(mfrow = c(2, 2))
plot(fit_full)
par(mfrow = c(1, 1))
```

Here are some important info we get. 

- $R_{adj}^2\approx0.74$ and $R^2\approx0.75$. So about $74-75\%$ of the variability in median house value is explained by the predictors.
- Overall p-value for the overall F-test is very small, which means that the whole group of variables work.
- `indus`, `age` and some of the `rad` levels are not signicant.

Then immediately do residual diagnostics on this full model:


The four diagnostic plots give me:

1. Residuals vs Fitted — checks linearity and homoscedasticity
2. Normal Q-Q — checks normality of residuals
3. Scale-Location — another view of heteroscedasticity
4. Residuals vs Leverage — identifies influential points (Cook's distance)

On raw Boston data, these plots typically show a fan shape (heteroscedasticity) and non-linearity — which tells me transformations are needed before I proceed.

### Variable selections

We apply best-of-subset to select variables.


```{r}
library(leaps)
fits <- regsubsets(
    medv ~ .,
    data = Boston,
    nvmax = ncol(model.matrix(medv ~ ., Boston)) - 1
)
models_summary <- summary(fits)
best_cp <- which.min(models_summary$cp)
plot(models_summary$cp, xlab = "Number of varaiables", type = "l")
points(best_cp, models_summary$cp[best_cp], col = "red", cex = 2, pch = 20)

models_summary
```

Since for dummy variable `rad` we have to choose all or none, based on the best-of-subset results, we will exclude `indus` and `age` from our model. We will delete the two variables from `Boston` and name the dataset `dat`.


```{r}
dat <- subset(Boston, select = -c(indus, age))
```


Therefore our based model is 
```{r}
fit_base <- lm(medv~., data=dat)
summary(fit_base)
```

The corresponding residual plots are
```{r}
par(mfrow = c(2, 2))
plot(fit_base)
par(mfrow = c(1, 1))
```




## Phase 3: Consider Transformations





### Check if `log(medv)` improves things

Based on all plots, we may need to deal with several curvy patterns as well as the non-constant variance issue. Therefore we would like to start from a transformation for `medv`. I'd also use Box-Cox to formally choose the best power transformation for `medv`:

```{r}
boxcox(fit_base)  
```

Reading this specific plot

- The peak appears to be right around $\lambda\approx0$, and 
- the 95% CI dotted lines straddle 0 closely. They look to be roughly $(-0.1, 0.2)$ or so.

By the standard rule, if $\lambda=0$ is inside the 95% CI, `log(y)` is the transformation.

```{r}
fit_log <- lm(log(medv) ~ ., data = dat)
par(mfrow = c(2, 2))
plot(fit_log)
par(mfrow = c(1, 1))
```

```{r}
summary(fit_log)
```

With `log(medv)` as the response:

- Coefficients are now interpreted as percent changes
- The residual fan shape (heteroscedasticity) is slightly improved


### Partial residual plots

Then we look at the partial residual plots.

First show the partial residual plots.

```{r}
par(mfrow = c(4, 3))
crPlots(fit_log)
par(mfrow = c(1, 1))
```

Based on the plots, it seems that `crim`, `rm`, `dis` and `lstat` is a little curvy. Therefore we would like to try squared terms for them.


```{r}
fit2 <- lm(log(medv)~.+I(crim^2)+I(rm^2)+I(dis^2)+I(lstat^2), data=dat)
summary(fit2)
par(mfrow = c(4, 3))
crPlots(fit2)
par(mfrow = c(1, 1))
```
