# Example: Boston Housing dataset full walkthrough

In the previous section we talk about our main tools to handle the `Boston` dataset. In this section we display how to explore the dataset from scratch.



| Variable  | Description                                                                 | Type                    |
|-----------|-----------------------------------------------------------------------------|-------------------------|
| `crim`    | Per capita crime rate by town                                                | Numerical, continuous   |
| `zn`      | Proportion of residential land zoned for lots over 25,000 sq.ft.            | Numerical, continuous   |
| `indus`   | Proportion of non-retail business acres per town                             | Numerical, continuous   |
| `chas`    | Charles River dummy variable (1 if tract bounds river; 0 otherwise)          | Categorical, nominal   |
| `nox`     | Nitrogen oxides concentration (parts per 10 million)                        | Numerical, continuous   |
| `rm`      | Average number of rooms per dwelling                                         | Numerical, continuous   |
| `age`     | Proportion of owner-occupied units built prior to 1940                       | Numerical, continuous   |
| `dis`     | Weighted mean distance to five Boston employment centers                     | Numerical, continuous   |
| `rad`     | Index of accessibility to radial highways (larger = better accessibility)   | Categorical, ordinal   |
| `tax`     | Full-value property-tax rate per \$10,000                                    | Numerical, continuous   |
| `ptratio` | Pupil–teacher ratio by town                                                  | Numerical, continuous   |
| `black`   | $1000(B_k - 0.63)^2$, where $B_k$ is the proportion of Black residents   | Numerical, continuous   |
| `lstat`   | Percentage of lower-status population                                       | Numerical, continuous   |
| `medv`    | Median value of owner-occupied homes (in \$1,000s)                          | Numerical, continuous   |



## Step 1: Understand the Data
Before touching `lm()`, the first step is always exploratory data analysis (EDA). 


```{r}
#| warning: false
library(MASS)
library(car)
data(Boston)

Boston$chas <- factor(Boston$chas)
Boston$rad <- factor(Boston$rad)

num_vars <- Boston[, sapply(Boston, is.numeric)]

summary(Boston)
```



```{r}
#| fig-height: 8
#| fig-show: hold

par(mfrow = c(4, 3))

for (n in names(num_vars)) {
    hist(num_vars[[n]], main = "", xlab = n)
}

par(mfrow = c(1, 1))
```


```{r}
pairs(Boston)
```

Correlation among numeric variables

```{r}
#| warning: false
cor_matrix <- cor(num_vars)

library(ggcorrplot)
ggcorrplot(cor_matrix, hc.order = TRUE)
```


Based on the summary table, histograms, pair plots, and the correlation matrix, here are the main takeaways from the EDA:

- The response `medv` is not perfectly symmetric and shows some upper-end compression, so a transformation of the response may help stabilize variance.
- `crim`, `zn`, `dis`, `black`, and `tax` show strong skewness or long tails in the summary table, with `crim` especially right-skewed. This suggests that transformations such as `log(crim)` may be useful.
- Several predictors are on very different scales, so raw coefficient sizes from a later regression should not be compared directly without standardization or context.
- The correlation matrix confirms that `medv` is most strongly negatively correlated with `lstat` and positively correlated with `rm`, which matches what we see in the pair plots.
- The pair plots show a strong negative association between `lstat` and `medv`, but the pattern is clearly curved rather than purely linear. This is a strong sign that a quadratic term or transformation will be needed.
- `rm` has a strong positive relationship with `medv`, and this is one of the clearest approximately linear patterns in the plots, although the spread increases somewhat for larger fitted values.
<!-- - `ptratio` tends to be negatively associated with `medv`, while `black` tends to be positively associated with `medv`, though both relationships appear weaker than those for `lstat` and `rm`. -->
- `nox`, `indus`, `age` and `dis` are highly correlated, positive or negative. These variables all describe urban intensity or location-related structure, so multicollinearity is a concern.
<!-- - `rad` and `tax` are strongly positively associated, indicating that highway accessibility and tax rate are carrying overlapping information. -->
<!-- - The matrix also confirms notable predictor-predictor correlations such as `nox` with `indus` and `age`, `dis` with `nox` in the opposite direction, and `rad` with `tax`, all of which warn us that the full model may suffer from collinearity. -->
- A few scatterplots suggest outliers or high-leverage towns, especially for `crim` and `tax`, so influence diagnostics will matter later even if we keep all observations.


## Phase 2: Start with All Predictors — Full Model

We begin with the full additive regression model.

```{r}
fit_full <- lm(medv ~ ., data = Boston)
summary(fit_full)
```
This model provides a baseline for comparison. 

```{r}
#| fig-show: hold
par(mfrow = c(2, 2))
plot(fit_full)
par(mfrow = c(1, 1))
```

Here are some important info we get. 

- $R_{adj}^2\approx0.74$ and $R^2\approx0.75$. So about $74-75\%$ of the variability in median house value is explained by the predictors.
- Overall p-value for the overall F-test is very small, which means that the whole group of variables work.
- `indus`, `age` and some of the `rad` levels are not signicant.

Then immediately do residual diagnostics on this full model:

<!-- 
The four diagnostic plots give me:

1. Residuals vs Fitted — checks linearity and homoscedasticity
2. Normal Q-Q — checks normality of residuals
3. Scale-Location — another view of heteroscedasticity
4. Residuals vs Leverage — identifies influential points (Cook's distance) -->

1. From the residual plot, the red line is not flat, with a slight U-shape, and hte residual become slightly larger for larger fitted value.
2. From the Q-Q plot, points follow the line in the middle, but right tail bends upward, and several points deviate strongly.
3. From the scale-location plot, the red trend line has a slight U-shape, and the residual spread grows as fitted values increase.
4. From the residuals vs leverage plot, most points have low leverage with one is around 0.30, and no points have large Cook's distance.


These suggest the following observations:

1. There are possible nonlinearity among variables, that we need to try some quadratic term or transformations.
2. There is a possible heteroscedasticity.
3. Residuals are not perfectly normal especially in the right tail.
4. There are no extremely influential observations.

<!-- On raw Boston data, these plots typically show a fan shape (heteroscedasticity) and non-linearity — which tells me transformations are needed before I proceed.
 -->
Based on all the observations (from this section and the previous section), we would like to start from applying some nonlinear transformations to build the model.



## Phase 3: Consider Transformations



### Check if `log(medv)` improves things


<!-- Based on all plots, we may need to deal with several curvy patterns as well as the non-constant variance issue. Therefore we would like to start from a transformation for `medv`. I'd  -->

We begin by examining the response variable. To determine an appropriate transformation, we apply the Box–Cox procedure to search for a suitable power transformation of `medv`:

```{r}
boxcox(fit_full)
```

The peak of the profile log-likelihood appears to occur around $\lambda\approx0.1$. The $95\%$ confidence interval for $\lambda$ is slightly above $0$, so $0$ lies just outside the interval. This suggests that the optimal transformation is approximately $y^{0.1}$. However, because $0.1$ is very close to $0$, the log transformation $\log(y)$ is often used in practice. The log transformation is simpler, widely used, and produces a model fit very similar to the optimal Box–Cox transformation.

We therefore refit the model using `log(medv)` as the response.
```{r}
#| fig-show: hold
fit_log <- lm(I(log(medv)) ~ ., data = Boston)
par(mfrow = c(2, 2))
plot(fit_log)
par(mfrow = c(1, 1))
```

```{r}
summary(fit_log)
```




With `log(medv)` as the response:

- Coefficients are now interpreted as percent changes in `medv` for a non-unit increase in the predictor, holding other variables constant.
- $R_a^2$ and $R^2$ increase slightly, indicating a small improvement in model fit.
- However, the diagnostic plots show that the main issues (nonlinearity, heteroscedasticity, and non-normal residuals) are not substantially improved.

Therefore, transforming the response alone is not sufficient, and we need to consider additional model modifications, such as transformations of predictors, interaction terms, or nonlinear terms.


### Partial residual plots

Then we look at the partial residual plots.

First show the partial residual plots.

```{r}
#| fig-show: hold
#| fig-width: 8
par(mfrow = c(4, 3))
crPlots(fit_log)
par(mfrow = c(1, 1))
```

Based on the plots, it seems that `crim`, `rm`, `dis` and `lstat` is a little curvy. Therefore we would like to try squared terms for them.


```{r}
#| fig-show: hold
fit2 <- lm(
    log(medv) ~ . + I(crim^2) + I(rm^2) + I(dis^2) + I(lstat^2),
    data = Boston
)
summary(fit2)
par(mfrow = c(4, 3))
crPlots(fit2)
par(mfrow = c(1, 1))
```

The quadratic terms improve the model substantially. Compared with `fit_log`, the residual standard error drops from about `0.189` to about `0.171`, and the nested F-test for adding the four squared terms is highly significant.

```{r}
anova(fit_log, fit2)
```

At this point, `zn` is no longer significant after the nonlinear terms are added, so we can simplify the model by dropping it and refitting.

```{r}
fit3 <- update(fit2, . ~ . - zn)
summary(fit3)
```

This simpler model has essentially the same fit as `fit2`, so removing `zn` costs us almost nothing while making the model easier to explain.

```{r}
anova(fit3, fit2)
```




### Variable selections

We apply best-of-subset to select variables.


```{r}
library(leaps)
fits <- regsubsets(
    medv ~ .,
    data = Boston,
    nvmax = ncol(model.matrix(medv ~ ., Boston)) - 1
)
models_summary <- summary(fits)
best_cp <- which.min(models_summary$cp)
plot(models_summary$cp, xlab = "Number of varaiables", type = "l")
points(best_cp, models_summary$cp[best_cp], col = "red", cex = 2, pch = 20)

models_summary
```

Since for dummy variable `rad` we have to choose all or none, based on the best-of-subset results, we will exclude `indus` and `age` from our model. We will delete the two variables from `Boston` and name the dataset `dat`.


```{r}
dat <- subset(Boston, select = -c(indus, age))
```


Therefore our based model is 
```{r}
fit_base <- lm(medv ~ ., data = dat)
summary(fit_base)
```

The corresponding residual plots are
```{r}
par(mfrow = c(2, 2))
plot(fit_base)
par(mfrow = c(1, 1))
```








## Phase 4: Diagnose the Final Model

Now we return to the standard diagnostics for the refined model.

```{r}
#| fig-show: hold
par(mfrow = c(2, 2))
plot(fit3)
par(mfrow = c(1, 1))
```

These plots are much better than the ones from the original untransformed model:

- The residuals versus fitted plot shows less curvature.
- The spread is more stable across fitted values.
- The Q-Q plot is closer to a straight line, although the tails still deviate somewhat.
- A few observations remain influential, which is common in this dataset.

We can also check multicollinearity and influence numerically.

```{r}
vif(fit3)
```

The polynomial terms create some moderate to large VIF values, especially for `rm` with `I(rm^2)` and `dis` with `I(dis^2)`. This is expected when a variable and its square are both in the model, so it is not automatically a reason to remove them.

```{r}
rs <- rstudent(fit3)
cd <- cooks.distance(fit3)
lev <- hatvalues(fit3)

c(
    max_abs_studentized = max(abs(rs)),
    num_abs_studentized_gt_3 = sum(abs(rs) > 3),
    num_cooks_gt_4_over_n = sum(cd > 4 / nrow(dat)),
    num_high_leverage = sum(lev > 2 * length(coef(fit3)) / nrow(dat))
)
```

This tells us that there are still some unusual observations, but the overall model is far more stable than the original additive fit on the raw response. Unless those data points are recording errors, I would usually keep them and report their presence rather than delete them automatically.

## Final Model

Our final working model is

$$
\log(\texttt{medv}) = \beta_0 + \beta_1\texttt{crim} + \beta_2\texttt{chas}
+ \beta_3\texttt{nox} + \beta_4\texttt{rm} + \beta_5\texttt{dis}
+ \beta_6\texttt{rad} + \beta_7\texttt{tax} + \beta_8\texttt{ptratio}
+ \beta_9\texttt{black} + \beta_{10}\texttt{lstat}
+ \beta_{11}\texttt{crim}^2 + \beta_{12}\texttt{rm}^2
+ \beta_{13}\texttt{dis}^2 + \beta_{14}\texttt{lstat}^2 + \varepsilon.
$$

In R formula form, this is:

```{r}
formula(fit3)
```

The main conclusions are:

- Transforming the response with `log(medv)` is justified by the Box-Cox plot and improves variance stability.
- Adding quadratic terms for `crim`, `rm`, `dis`, and `lstat` captures important curvature that the additive linear model misses.
- After those nonlinear terms are included, `zn` is no longer needed.
- The final model explains about `83%` of the variation in `log(medv)`, with adjusted $R^2$ around `0.825`.

## Takeaway

This example shows a typical regression workflow:

1. Start with EDA and a full model.
2. Use residual plots and partial residual plots to detect problems.
3. Transform the response when the variance and shape suggest it.
4. Add nonlinear terms where the data indicate curvature.
5. Recheck diagnostics before settling on the final model.

The important point is that model building is iterative. We do not fit one regression and stop; instead, we repeatedly use diagnostics to decide how to improve the model.
