# Residual analysis

{{< include ../math.qmd >}} 

Residual analysis is one of the main tools for checking whether a regression model is appropriate.  
In a regression model, the random errors are not directly observable, so we study the **residuals** instead.

Recall the usual assumptions on the error term $\varepsilon$ in a regression model:

1. $\varepsilon$ is normally distributed.
2. $\varepsilon$ has mean $0$.
3. $\varepsilon$ has constant variance $\sigma^2$.
4. All pairs of error terms are uncorrelated.

 
This chapter introduces graphical tools and statistical ideas for detecting important departures from these assumptions.

## Regression Residuals

Suppose the regression model is
$$
y_i = \beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip} + \varepsilon_i.
$$

Because the true error $\varepsilon_i$ is unobservable, we estimate it by the residual
$$
\hat \varepsilon_i = y_i - \hat y_i,
$$
where $\hat y_i$ is the fitted value from the regression model.

Therefore a residual measures how far the observed response is from the fitted response.

- If $e_i>0$, the model underpredicts the response at observation $i$.
- If $e_i<0$, the model overpredicts the response at observation $i$.
- Large residuals indicate observations that are not well explained by the fitted model.


::: {#prp-}
## Two useful properties of residuals

For an ordinary least squares regression model with an intercept:

1. The residuals sum to zero:
   $$
   \sum_{i=1}^n e_i = 0.
   $$

2. The standard deviation of the residuals is equal to the standard deviation of the fitted regression model, $s$:
   $$
    s=\sqrt{\frac{SSE}{n-(p+1)}}
   $$

:::

## Graphical Displays of Residuals

Residual plots are used to assess whether the model assumptions appear reasonable.

If the assumptions are satisfied, then residual plots should generally show:

- no systematic trend,
- no obvious curvature,
- no dramatic increase or decrease in spread,
- only a few residuals (around $5\%$) more than 2 estimated standard deviations ($2s$) of $\varepsilon$ above or below $0$.

A good residual plot looks like a random cloud centered around $0$.

Residual plots are often the first diagnostic tool used to determine whether the deterministic part of the model may be misspecified.


::: {#prp-}
## Detecting Model Lack of Fit with Residuals

1. Plot the residuals, $\hat\varepsilon$, on the vertical axis against each of the independent variables $x_1$, ..., $x_k$ on the horizontal axis.
2. Plot the residuals, $\hat\varepsilon$, on the vertical axis against the predicted value $\hat y$ on the horizontal axis.
3. In each plot, look for 
   - systematic treands,
   - curvature,
   - dramatic changes in variability,
   - or more than $5\%$ of residuals outside the range $\pm 2s$.

Any of these patterns suggests that the model may not fit the data adequately.
:::




::: {#exm-}
## Detecting Curvature

<details>
<summary><strong>Click to expand.</strong></summary>



```{r}
set.seed(123)

n <- 500

x <- runif(n, 12, 30)
y <- 55.7 + 0.5969 * x - 2 * (x)^2 + rnorm(n, 0, 10)
```

Linear model

```{r}
fit <- lm(y ~ x)
plot(x, resid(fit), xlab = "x", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
```

Quadratic model

```{r}
fit2 <- lm(y ~ x + I(x^2))
plot(x, resid(fit2), xlab = "x", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
```

The residual plot from the linear model typically shows a parabolic pattern, indicating that the straight-line model is inadequate. After including the quadratic term, the residuals become more randomly scattered, suggesting the quadratic model better captures the mean structure.


</details>

:::



::: {#exm-}
## Multiple Predictors

<details>
<summary><strong>Click to expand.</strong></summary>



```{r}
set.seed(123)

n <- 500

x1 <- runif(n, 10, 20)
x2 <- runif(n, 15, 30)
y <- 5 + 4 * x1 + 3 * x2^2 + rnorm(n, 0, 5)
```

Linear model

```{r}
#| fig-show: hold
#| fig-width: 10
#| fig-height: 4
fit <- lm(y ~ x1 + x2)

par(mfrow = c(1, 3))
plot(x1, resid(fit), xlab = "x1", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
plot(x2, resid(fit), xlab = "x2", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
plot(y, resid(fit), xlab = "fitted y", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
par(mfrow = c(1, 1))
```


Model with quadratic term


```{r}
#| fig-show: hold
#| fig-width: 10
#| fig-height: 4
fit2 <- lm(y ~ x1 + x2 + I(x2^2))

par(mfrow = c(1, 3))
plot(x1, resid(fit2), xlab = "x1", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
plot(x2, resid(fit2), xlab = "x2", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
plot(y, resid(fit2), xlab = "y", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
par(mfrow = c(1, 1))
```




</details>

:::



When drawing the plot that fitted values vs residuals, you may manually plot it as shown in the examples, or you may use `plot(fit, which=1)`. This `plot(fit)` will be discussed later.


<!-- 
### Typical interpretation

- A **random cloud** of residuals suggests the fitted mean structure is reasonable.
- A **parabolic pattern** suggests a quadratic term may be needed.
- A **nonlinear pattern** suggests transformation or a more flexible model may be appropriate.

### Example idea from the lecture

For the cholesterol-versus-fat-intake data, the first-order model produced a residual plot with a parabolic pattern. That suggests the straight-line model is inadequate and that a second-order model may fit better. After fitting the quadratic model, the residual plot no longer showed a clear pattern, so the quadratic model was more appropriate. -->

### Partial Residual Plots

A partial residual plot (also called a component + residual plot) helps assess the relationship between the response and one predictor after accounting for the other predictors in the model.

For predictor $x_j$, the partial residual is 
$$
e_i + \hat\beta_j x_{ij}.
$$

Plotting these values against $x_j$ can reveal structure that may be hidden in an ordinary residual plot. Such plots can help detect:

- curvature in the relationship between the response and a predictor,
- the need for transformation of a predictor,
- nonlinear relationships not visible in standard residual plots.



A partial residual plot may suggest:

- adding a quadratic term such as $x_j^2$,
- transforming the variable, for example $\log(x_j)$ or $1/x_j$,
- reconsidering the functional form of the model.

The function `car::crPlots()` produces partial residual plots for each predictor. Each plot contains:

- Points: component + residual values
- Blue line: the fitted linear component $\hat\beta_j x_j$
- Red line: a smooth trend (typically obtained using LOESS smoothing) fitted to the partial residuals



If the red and blue lines are similar, the linear specification for that predictor is reasonable. If the red line deviates noticeably, especially if it shows curvature, the model may need a transformation or nonlinear term.
<!-- 
We may use `car::crPlots` to make the parttial residual plot for each variable. In the plot, other than the points representing the component+residuals, there is a red line and a blue line. The blue line is the straight line based on $\hat\beta_jx_j$, and the red line is a smooth trend line, typically obtained using LOESS smoothing, applied to the partial residuals. 

- If the two lines match, this variable is explained by the model well. 
- If the two lines differ, especially if the red line is visually curved, this variable is not represented by the model well, and it is highly possible that a nonlinear transformation should be applied to this variable. -->


::: {#exm-}


```{r}
set.seed(123)

n <- 500

x1 <- runif(n, .1, 1)
x2 <- runif(n, 0, 2)
y <- 5 + 1 / x1 + 3 * x2^2 + rnorm(n, 0, 1)
```


```{r}
#| warning: false
#| fig-width: 10
#| fig-height: 4
library(car)

fit <- lm(y ~ x1 + x2)
summary(fit)
crPlots(fit)
```

For both predictors, the red smooth curves show clear curvature, suggesting that the linear specification is inadequate.

Based on the shapes of the curves:

- for $x_1$, a reciprocal transformation appears appropriate,
- for $x_2$, a quadratic term appears appropriate.



```{r}
#| fig-width: 10
#| fig-height: 4
fit1 <- lm(y ~ I(1 / x1) + I(x2^2))
summary(fit1)
crPlots(fit1)
```

After applying these transformations, the red and blue lines closely match, indicating that the functional form is now more appropriate.





:::




















## Detecting Unequal Variances

One important regression assumption is that the error variance is constant:
$$
\Var(\varepsilon_i)=\sigma^2 \quad \text{for all } i.
$$


::: {#def-}


- **Homoscedasticity**: constant variance.
- **Heteroscedasticity**: unequal variance.

:::

Heteroscedasticity can often be detected from a residual plot. A plot of residuals versus fitted values, or versus a predictor, may show

- a funnel shape,
- a cone shape,
- spread increasing with the mean,
- spread decreasing with the mean.

These patterns suggest that the variability of the response changes with the level of the predictor or fitted value. In many such cases, the variance of $y$ is a function of its mean $\Exp(y)$. 



::: {#exm-}
## Common reasons for heteroscedasticity

Heteroscedasticity often occurs when the variance of the response depends on its mean. Common situations include:

- count data, such as Poisson responses,


```{r}
set.seed(1)

n <- 2000
x <- runif(n, 0, 25)
y <- rpois(n, 1 + 2 * x)

dat1 <- data.frame(x = x, y = y)

fit1 <- lm(y ~ x, data = dat1)

plot(fitted(fit1), resid(fit1), xlab = "Fitted value", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
```


- proportions or percentages, such as binomial responses,


```{r}
set.seed(1)

n <- 2000
m <- 100
x <- runif(n, 0, 1)
y <- rbinom(n, m, x) / m

dat2 <- data.frame(x = x, y = y)

fit2 <- lm(y ~ x, data = dat2)

plot(fitted(fit2), resid(fit2), xlab = "Fitted value", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
```



- multiplicative error structures.

```{r}
set.seed(1)

n <- 2000
x <- runif(n, 1, 20)
eps <- exp(rnorm(n, 0, 0.4))
y <- x * eps

dat3 <- data.frame(x = x, y = y)

fit3 <- lm(y ~ x, data = dat3)

plot(fitted(fit3), resid(fit3), xlab = "Fitted value", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
```


:::




A common remedy is to apply a response transformation chosen to stabilize the variance.

Typical variance-stabilizing transformations include:

- $\sqrt{y}$ for count-type data,
- $\log(y)$ for multiplicative or right-skewed responses,
- other Box--Cox transformations: see @sec-boxcox. 

| Type of Response      | Variance                     | Stabilizing Transformation |
|:----------------------|:-----------------------------|:---------------------------|
| Poisson               | $\Var(y)=\Exp(y)$                       | $\sqrt{y}$                 |
| Binomial proportion   | $\Var(y)=\dfrac{\Exp(y)(1-\Exp(y))}{m}$    | $\sin^{-1}(\sqrt{y})$      |
| Multiplicative        | $\Var(y)\propto [\Exp(y)]^2$          | $\log(y)$                   |

: Stabilizing transformations for heteroscedastic responses          


::: {#exm-}
## Count data

For count data, a square-root transformation is often used to stabilize the variance.


```{r}
set.seed(1)

n <- 2000
x <- runif(n, 0, 25)
y <- rpois(n, 1 + 2 * x)

dat1 <- data.frame(x = x, y = y)

fit1 <- lm(I(sqrt(y)) ~ x, data = dat1)

plot(fitted(fit1), resid(fit1), xlab = "Fitted value", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
```



<details>
<summary><strong>Click to expand.</strong></summary>

Note that since there is an obvious curve, we could consider to add $x^2$ term.


```{r}
fit1c <- lm(I(sqrt(y)) ~ x + I(x^2), data = dat1)

plot(fitted(fit1c), resid(fit1c), xlab = "Fitted value", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
summary(fit1c)
```



Here are some calculations related to the model. Suppose the response follows a Poisson model

$$
Y \sim \text{Poisson}(\lambda), \qquad \lambda = 1 + 2x .
$$

We consider the transformed response $\sqrt{Y}$. Using a second–order delta method approximation,

$$
E(\sqrt{Y}) \approx \sqrt{\lambda} - \frac{1}{8\sqrt{\lambda}},
\qquad \lambda = 1 + 2x .
$$

Let
$$
m(x) = \sqrt{1+2x} - \frac{1}{8\sqrt{1+2x}}.
$$

We approximate $m(x)$ near $x_0 = 12.5$ using the 2nd order Taylor polynomial. We choose $12.5$ since $12.5$ is the mean of $\distunif(0,25)$. At this point,

$$
\lambda_0 = 1 + 2x_0 = 26 .
$$

Evaluating derivatives 
$$
\begin{split}
m'(x)&=\frac1{\sqrt{1+2x}}+\frac{1}{8(1+2x)^{3/2}},\\
m''(x)&=-\frac{1}{(1+2x)^{3/2}}-\frac3{8(1+2x)^{5/2}}.
\end{split}
$$

Then

$$
m(12.5) \approx 5.0745, \qquad
m'(12.5) \approx 0.1971, \qquad
m''(12.5) \approx -0.00765 .
$$

The second-order Taylor expansion around $12.5$ is

$$
m(x)
\approx
5.0745
+
0.1971(x-12.5)
-
0.00382(x-12.5)^2 \approx 2.014 + 0.293x - 0.00382x^2 .
$$

Compare it to our estimation.

```{r}
fit1c
```



</details>



:::

::: {#exm-}
## Binomial-type


For binomial-type responses, the arcsine square-root transformation is sometimes used to reduce heteroscedasticity.


```{r}
set.seed(1)

n <- 2000
m <- 100
x <- runif(n, 0, 1)
y <- rbinom(n, m, x) / m

dat2 <- data.frame(x = x, y = y)

fit2 <- lm(I(asin(sqrt(y))) ~ x, data = dat2)

plot(fitted(fit2), resid(fit2), xlab = "Fitted value", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
```


<details>
<summary><strong>Click to expand.</strong></summary>
Similar to the previous example, we could add 2nd order and 3rd order term since there is an obvious curvy pattern.


```{r}
fit2c <- lm(I(asin(sqrt(y))) ~ x + I(x^2) + I(x^3), data = dat2)

plot(fitted(fit2c), resid(fit2c), xlab = "Fitted value", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)

summary(fit2c)
```



</details>

:::

::: {#exm-}
## Multiplicative error structures

When the variability increases proportionally to the mean, a log transformation is often useful. A more natural multiplicative-error model is $y=x\cdot\varepsilon$, where $\varepsilon>0$ is a random error term. Taking logs gives
$$
\log y=\log x+\log \varepsilon,
$$
so the log transformation converts multiplicative errors into additive ones.

```{r}
set.seed(1)

n <- 2000
x <- runif(n, 1, 20)
eps <- exp(rnorm(n, 0, 0.4))
y <- x * eps

dat3 <- data.frame(x = x, y = y)

fit3 <- lm(log(y) ~ log(x), data = dat3)

plot(fitted(fit3), resid(fit3), xlab = "Fitted value", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
```


:::


::: {.callout-tip}
For count and proportion responses, generalized linear models (GLM) are often more natural than transforming the response and fitting ordinary least squares.
:::


### Box-Cox transformation {#sec-boxcox}


When a simple transformation such as $\sqrt{y}$ or $\log(y)$ is not clearly appropriate, the Box-Cox transformation provides a systematic family of power transformations for a positive response variable $y>0$.

For a parameter $\lambda$, the Box--Cox transformation is defined as
$$
y^{(\lambda)}=
\begin{cases}
\dfrac{y^\lambda-1}{\lambda}, & \lambda \neq 0,\\[1em]
\log(y), & \lambda = 0.
\end{cases}
$$

This family includes several common transformations as special cases:

- $\lambda=1$: no transformation (approximately),
- $\lambda=\frac{1}{2}$: square-root transformation,
- $\lambda=0$: log transformation,
- $\lambda=-1$: reciprocal transformation.

<!-- The main idea is to choose a value of $\lambda$ that makes the regression assumptions more reasonable, especially:

- linearity,
- constant variance,
- approximate normality of errors. -->

In practice, $\lambda$ is usually estimated from the data by maximizing the likelihood over a range of candidate values.

A fitted model after transformation has the form
$$
y^{(\lambda)}=\beta_0+\beta_1x_1+\cdots+\beta_px_p+\varepsilon.
$$

After fitting the model, we check whether the transformed response leads to improved residual plots and more stable variance.

::: {.callout-warning}
## Important restriction

The Box--Cox transformation requires $y>0$. If the response contains zeros or negative values, this method cannot be applied directly without modification.
:::





::: {#exm-}
## Box-Cox transformation



<details>
<summary><strong>Click to expand.</strong></summary>


```{r}
set.seed(123)

n <- 500

x1 <- runif(n, 1, 2)
x2 <- runif(n, 1, 3)
y <- exp(3 * x1 + 4 * x2 + rnorm(n, 0, 1))
dat <- data.frame(x1 = x1, x2 = x2, y = y)
```

```{r}
library(MASS)

fit <- lm(y ~ x1 + x2, data = dat)
boxcox(fit)
```

The plot shows that $\lambda$ is very close to $0$. Therefore it is recommended to use `log(y)` as the transformation. You can compare the results below.

```{r}
fit1 <- lm(y ~ ., data = dat)
summary(fit1)
plot(fit1, which = 1)
```

```{r}
fit2 <- lm(I(sqrt(y)) ~ ., data = dat)
summary(fit2)
plot(fit2, which = 1)
```


```{r}
fit3 <- lm(I(log(y)) ~ ., data = dat)
summary(fit3)
plot(fit3, which = 1)
```


</details>



:::



<!-- 
Box--Cox is most useful when:

- residual plots suggest heteroscedasticity or nonlinearity,
- the response is positive and right-skewed,
- no obvious transformation is known in advance.

It is best viewed as a guide for choosing a reasonable transformation rather than a purely automatic rule. 

### Example idea from the lecture

For the social worker salary data, a second-order model appeared to fit the mean structure reasonably well, but the residual plot had a cone shape, indicating non-constant variance. After transforming the response to $\log(y)$, the cone shape was greatly reduced, and the model assumptions were more plausible. -->

<!-- ### A simple test for heteroscedasticity

One classical approach is to split the data into two groups, fit the model separately, and compare the two error variances using an $F$-test.

If the estimated variances differ substantially, this provides evidence against equal variances. -->

## Checking the Normality Assumption

Normality is another standard assumption used for many regression inference procedures.
<!-- 
### Graphical tools for checking normality -->

Common tools include:

- histogram of residuals,
- stem-and-leaf display,
- normal probability plot (QQ plot).

Among these, the QQ plot is often the most informative.


### Interpreting a normal probability plot {.unnumbered}

- The x-axis represents the residuals, and the y-axis represents the normal scores. 
- If the points follow an approximately straight line, the normality assumption is reasonable.
- Strong curvature or systematic deviation from a line suggests nonnormality.



::: {#exm-}
## Examples of QQ plots


```{r}
set.seed(1)

n <- 2000
x <- runif(n, 1, 2)
eps <- rnorm(n)
y <- 10 + 2 * x + eps
dat1 <- data.frame(x = x, y = y)
fit1 <- lm(y ~ x, data = dat1)

plot(fit1, which = 2)
```


:::
<!-- 
### Practical comment

Regression methods are often fairly robust to moderate nonnormality when the sample size is reasonably large. However, severe skewness or heavy tails may still affect inference. -->

<!-- ### Fixing nonnormality -->

Nonnormality is often accompanied by heteroscedasticity. In many cases, a transformation of the response helps with both problems.

A standard family of transformations is the Box--Cox transformation, which can help identify a suitable power transformation for the response.

::: {#exm-}
## `dat2`

```{r}
set.seed(1)

n <- 2000
x <- runif(n, 1, 2)
mu <- exp(1 + x)
y <- rlnorm(n, log(mu), 0.5)
dat2 <- data.frame(x = x, y = y)
fit2 <- lm(y ~ x, data = dat2)

plot(fit2, which = 2)
```

Since
```{r}
boxcox(fit2)
```

$\lambda=0$. Therefore we need to transform $y\mapsto \log(y)$.
```{r}
fit2c <- lm(I(log(y)) ~ x, data = dat2)
plot(fit2c, which = 2)
```

The non-normality is fixed by the Box-Cox transformation in this case.

:::



::: {#exm-}
## `dat3`
Consider `dat3`. 


```{r}
set.seed(1)

n <- 2000
x <- runif(n, 1, 2)
eps <- rnorm(n, 0, 0.25 * x)^3
y <- 5 + 2 * x + eps
dat3 <- data.frame(x = x, y = y)
fit3 <- lm(y ~ x, data = dat3)

plot(fit3, which = 2)
```

Since
```{r}
boxcox(fit3)
```

$\lambda\approx 0.5$. Therefore we need to transform $y\mapsto \sqrt{y}$.
```{r}
fit3c <- lm(I(sqrt(y)) ~ x, data = dat3)
plot(fit3c, which = 2)
```

Note that the non-normality is almost not fixed in this case.

:::



::: {.callout-caution}
Among the assumptions of the linear regression model, the normality assumption usually has the least impact on the estimation itself. The main role of the normality assumption is to justify the use of t-tests and F-tests for assessing the significance of variables.

If the error terms are not normally distributed, these tests may not be exact when the sample size is small. In such cases, directly applying the usual t-test and F-test formulas may lead to less accurate inference.

Therefore, it is desirable for the error term to be approximately normally distributed. However, in practice this assumption is not crucial. As long as the other key assumptions—such as linearity, independence, and constant variance—are reasonably satisfied, the regression results are usually still reliable, especially when the sample size is large. In large samples, the central limit theorem implies that the sampling distribution of the estimated coefficients is approximately normal, even if the error terms are not normally distributed.
:::


::: {.callout-note}
## Box-Cox transformation
The Box–Cox transformation is used to find a power transformation of the response variable that stabilizes variance and improves the linear model assumptions. So its main usage is to stabilize variance, and it could move the error term towards the normal distribution as a byproduct.
:::


## Detecting Outliers and Influential Observations

Not every unusual point has the same meaning. Some are merely outliers in the response direction, while others strongly affect the fitted model.

### Standardized residuals

A **standardized residual** rescales a residual by an estimate of its standard deviation.

A common rule of thumb is:

- $|r_i| > 2$ may indicate a possible outlier,
- $|r_i| > 3$ indicates a more serious outlier.

### Studentized deleted residuals

A **studentized deleted residual** is usually preferred because it measures how unusual a point is when that same point is omitted from the variance estimate.

A common rule of thumb is:

- values outside $[-2,2]$ suggest possible outliers.

### Important note

An outlier should not automatically be removed. We first ask:

- Is it a data-entry or recording error?
- Is the model missing important terms?
- Is the point valid but scientifically unusual?

Only obvious recording errors should be automatically corrected or removed.

## Influential Observations

An observation is **influential** if it has a substantial effect on the fitted model, parameter estimates, or conclusions.

### Leverage

Leverage measures how extreme an observation is in the predictor space.

High-leverage points have unusual predictor values and may strongly affect the fitted line or surface.

A common rule of thumb flags an observation if its leverage is larger than
$$
\frac{2(p+1)}{n}
$$
or sometimes
$$
\frac{3(p+1)}{n},
$$
where $p+1$ is the number of parameters in the model including the intercept.

### Cook's distance

**Cook's distance** combines information about residual size and leverage to assess influence.

Common guidelines:

- $D_i > 1$ suggests an influential observation,
- larger Cook's distance values deserve further investigation.

Cook's distance is often preferred because it captures overall impact on the fitted regression.

### Example idea from the lecture

In the fast-food sales example, one observation had an extremely large standardized residual. Investigation showed that the recorded sales value contained a data-entry error. After correcting the value, the fitted model improved dramatically, showing how a single bad data point can distort a regression analysis.

## Summary of Diagnostic Tools

### For lack of fit

Use:

- residuals versus predictor,
- residuals versus fitted values,
- partial residual plots.

Look for:

- curvature,
- trends,
- nonrandom structure.

### For unequal variance

Use:

- residuals versus fitted values,
- residuals versus predictors.

Look for:

- cone shape,
- funnel shape,
- changing spread.

### For normality

Use:

- histogram of residuals,
- normal probability plot.

Look for:

- approximate linear trend in the QQ plot.

### For outliers

Use:

- standardized residuals,
- studentized deleted residuals.

### For influential observations

Use:

- leverage,
- Cook's distance.

## Suggested Model-Building Procedure

A practical regression diagnostic workflow is:

1. Fit an initial model.
2. Examine residual plots for lack of fit.
3. Check for heteroscedasticity.
4. Check the normality of residuals.
5. Identify possible outliers and influential points.
6. Investigate unusual observations carefully.
7. Modify the model if needed:
   - add missing terms,
   - transform variables,
   - correct recording errors,
   - refit and reassess.

## Key Takeaways

- Residuals are proxies for the unobserved regression errors.
- Residual plots are central tools for diagnosing model adequacy.
- Curvature in a residual plot suggests lack of fit.
- Funnel or cone shapes suggest heteroscedasticity.
- QQ plots help assess normality.
- Standardized residuals and studentized deleted residuals help detect outliers.
- Leverage and Cook's distance help detect influential observations.
- Model revision should be guided by diagnostics and subject-matter reasoning, not by automatic deletion of unusual points.
