



# Variable selection


{{< include ../math.qmd >}} 

When we build a multiple regression model, we often begin with many possible predictors. The main question is: **which variables should be included in the model?**

This chapter introduces two systematic methods for screening variables:

1. **Stepwise regression**
2. **All-possible-regressions selection** (also called **best-subsets regression**)

These methods are intended to help identify the most important predictors before the more careful model-building phase begins.

<!-- ## Why Use a Variable Screening Method

In many applications, researchers collect a response variable $y$ together with a large number of potential independent variables $x_1, x_2, \dots, x_p$. Examples include predicting:

- firm profit,
- college GPA,
- inflation rate,
- salary,
- or other outcomes of interest.

When there are many candidate predictors, it can be difficult to decide which variables should enter the regression model. Variable screening methods provide a more systematic way to reduce the list.

The two methods emphasized in this chapter are:

- **Stepwise regression**
- **All-possible-regressions selection** -->

## Model Selection Criteria
Variable selections can be treated as a model comparison process. Therefore we need to discuss criteria on comparing models.

### What we already covered
1. p-value: when comparing nested models, the p-value can be used to determine whether the added variables are significant as a group.
2. $R^2$ and $R_a^2$: the proportion of the varaiance that can be explained. This is one of the metrics that is used to determine the performance of a model.
3. MSE: this measures the average squared prediction error of the model on the training data.

When talking about approximation ability, some tend to use MSE. This is very similar to $R^2$ and $R_a^2$ since they are computed using MSE.

### AIC
The Akaike Information Criterion (AIC) is a statistical metric used to evaluate the performance of a model. It balances model fit (goodness of fit) with complexity to avoid over or underfitting. Lower AIC values indicate a better, more efficient model.

A common formula is

$$
AIC = -2 L + 2k,
$$

where $L$ is the likelihood of the fitted model and $k$ is the number of estimated parameters. The first term rewards good fit, while the second term  penalizes model complexity. In the case of linear regression, AIC is 
$$
AIC=n\log\frac{SSE}{n}+2k
$$
where $SSE$ is the sum of squared errors of the model, $n$ is the number of observations and $k$ is the number of estimated predictors. 

> AIC can be treated as adjusting the training error with a penalty term to provide an unbiased estimate of the test error.



::: {.callout-note collapse="true"}
## The rough story of AIC

AIC is derived as an estimate of expected out-of-sample prediction error, measured via Kullback–Leibler (KL) divergence.

1. Let $g(y)$ be the true but unknown data-generating distribution coming from the real world, and $f(y\mid\theta)$ be the candidate model to estimate $g(y)$, where $\theta$ is the parameters vector of the model. 
2. Akaike's idea to measure the difference between $g$ and $f$ is KL divergence $D_{KL}$. We would like to find the parameters $\hat\theta$ such that $D_{KL}(g\,\|\,f_{\theta})$ is minimized.
$$
\begin{split}
D_{KL}(g\,\|\,f)&=\int g(y)\log\frac{g(y)}{f(y\mid \theta)}\dl3y=\underbrace{\int g(y)\log g(y)\dl3y}_{\text{constant}}-\int g(y)\log f(y\mid\theta)\dl3y\\
&=\int g(y)\log g(y)\dl3y-\Exp_g[\log f(Y\mid\theta)].
\end{split}
$$
So we need to find the $\hat\theta$ such that $\Exp_g[\log f(Y\mid\theta)]$ is maximized.
3. Since we don't know the real $g$, but we have samples from real world $Y_1,\ldots,Y_n$, so we maximize the sample log-likelihood:
$$
\ell(\theta)=\sum_{i=1}^n\log f(Y_i\mid\theta).
$$
This gives the MLE estimator $\hat\theta$.
4. However $\ell(\hat\theta)$ is optimistically biased for the out-of-sample log-likelihood $\ell^*(\hat\theta)$, because the same data is used for fitting and evaluation.
5. Akaike’s key result (asymptotic bias correction) proved that under some regularity conditions, 
$$
\Exp_g[\ell(\hat\theta)]\approx \Exp_g[\ell^*(\hat\theta)]+k,
$$
where $k$ is the number of estimated parameters. Therefore 
$$
\Exp_g[\ell^*(\hat\theta)]\approx \Exp_g[\ell(\hat\theta)-k].
$$
So a natural estimator for $\ell^*(\hat\theta)$ is $\ell(\hat\theta)-k$. 
6. The estimator suggests the definition
$$
AIC=-2\ell(\hat\theta)+2k.
$$
The factor $-2$ is conventional. It puts AIC on the deviance scale, making it consistent with likelihood-ratio theory and convenient for model comparison.
7. Consider a MLR model, $Y=X^T\beta+\varepsilon$, where $\varepsilon\sim N(0,\sigma^2)$. Then based on the MLE estimator calculation, the log-likelihood is
$$
\ell(\beta,\sigma^2)= -\frac{n}{2}\log(2\pi)- \frac{n}{2}\log(\sigma^2)- \frac{1}{2\sigma^2}\mathrm{SSE}.
$$
Using the MLE $\hat{\sigma}^2 = \mathrm{SSE}/n$, we obtain
$$
\ell(\hat{\theta})= -\frac{n}{2}\left[\log(2\pi) + 1 + \log\left(\frac{\mathrm{SSE}}{n}\right)\right]=-\frac{n}{2}\log\left(\frac{\mathrm{SSE}}{n}\right)- \underbrace{\frac{n}{2}\left(1 + \log(2\pi)\right)}_{\text{constant}}.
$$
8. Multiplying by $-2$ and add $2k$ gives the final form of AIC in MLR:
$$
AIC=n\log\qty(\frac{SSE}n)+2k+\underbrace{n\log(2\pi)}_{\text{constant}}.
$$
9. Since the constant does not depend on the model, for comparing linear regression models we might abuse the notation and use
$$
AIC=n\log\qty(\frac{SSE}n)+2k.
$$

:::


::: {.callout-caution}
## About $k$
The $k$ is AIC is the number of estiamted parameters. In the case of MLR, all parameters that have to be estimated include all $\beta$'s as well as the variance of the error term $\sigma^2$.

- $\beta_0$: 1
- One $\beta_i$ for each predicator: $p$
- $\sigma^2$: 1

Therefore if there are $p$ predicator, $k$ is usually $p+2$. 
:::

::: {#exm-}

```{r}
set.seed(123)
n <- 30
x1 <- runif(n, 1, 10)
x2 <- runif(n, 1, 10)
y <- 20 + 5 * x1 - 4 * x2 + rnorm(n, 0, 5)
fit <- lm(y ~ x1 + x2)
AIC(fit)
```

We can compare it with the manual calculation.

```{r}
sse <- anova(fit)["Residuals", "Sum Sq"]
p <- 2
k <- p + 2
aic_manual <- n * log(sse / n) + n * (1 + log(2 * pi)) + 2 * k
aic_manual
```

:::





Based on the formula
$$
AIC=n\log\qty(\frac{SSE}n)+2k+C,
$$
there might be some variations. Consider the Taylor polynomial expansion of $\log x$ at $x_0$:
$$
\log x\approx \log x_0+\frac1{x_0}(x-x_0)=\log x_0-1+\frac x{x_0}.
$$
Let $x_0=\hat\sigma^2$. We have
$$
\begin{split}
\frac{AIC}n&\sim \log\qty(\frac{SSE}n)+\frac{2k}n\\
&\approx \log \hat\sigma^2-1+\frac1{\hat\sigma^2}\frac{SSE}{n}+\frac{2k}n\\
&=\frac{1}{\hat\sigma^2}\frac1n\qty(SSE+2k\hat\sigma^2)+C.
\end{split}
$$
Therefore if we ignore the added constant
$$
\frac{\hat\sigma^2}n AIC\approx \frac1n \qty(SSE + 2k\hat\sigma^2).
$$

When we treat both $\hat\sigma^2$ and $n$ as constants that are approximately irrevalent to the model, we have may have the relation
$$
AIC\propto \frac1n \qty(SSE + 2k\hat\sigma^2).
$$ {#eq-aic}
This is the definition of AIC in some references @Jam2021. 





### Mallows’ $C_p$

Mallows' $C_p$ is a criterion for comparing subset regression models. It balances goodness of fit and model complexity, with the goal of estimating prediction error. For a candidate subset model with $p$ parameters (including the intercept), Mallows' $C_p$ is defined as 
$$
C_p=\frac{SSE_p}{\hat\sigma^2}-(n-2p),
$$

where:

- $SSE_p$ is the residual sum of squares for the subset model,
- $\hat\sigma^2=MSE_{\text{full}}$ is the mean square error estimated from the full model,
- $n$ is the sample size,
- $p$ is the number of parameters in the subset model.



How to interpret it:

- A smaller $C_p$ is better.
- A good model typically has $C_p\approx p$.
- If $C_p$ is much larger than $p$, the model likely suffers from bias that important variables are missing.
- If $C_p<p$, it may indicate overfitting or that $\hat\sigma^2$ is overestimated.


::: {#exm-}

- Model A has $p=3$ parameters and $C_p=12$
- Model B has $p=5$ parameters and $C_p=5.4$

Model B is usually preferred. Its $C_p$ is small and very close to $p=5$, which suggests that the model is capturing the important predictors without too much bias. Model A has a much larger $C_p$ than $p=3$, so it likely leaves out useful variables.
:::


<!-- ::: {.callout-note collapse="true"}
## A rough story of $C_p$



::: -->


::: {#exm-}

```{r}
set.seed(123)
n <- 30
x1 <- runif(n, 1, 10)
x2 <- runif(n, 1, 10)
y <- 20 + 5 * x1 - 4 * x2 + rnorm(n, 0, 5)
fit <- lm(y ~ x1 + x2)
```

```{r}
fit_full <- lm(y ~ x1 + x2)
sigma2_hat <- summary(fit_full)$sigma^2 # MSE_full

fit_sub <- lm(y ~ x1)

SSE_p <- sum(residuals(fit_sub)^2)
p <- length(coef(fit_sub))

Cp <- SSE_p / sigma2_hat - (n - 2 * p)
Cp
```

:::

$C_p$ also has other formats. Rewrite its formula. Since
$$
\frac1n\hat\sigma^2(C_p+n)=\frac1n\hat\sigma^2\qty(\frac{SSE_p}{\hat\sigma^2}-n+2p)=\frac{1}n\qty(SSE_p+2p\hat\sigma^2),
$$
this shows that $C_p$ corresponds to 
$$
\text{training error + penality for model size.}
$$
Therefore some reference uses this formula to define $C_p$ @Jam2021:
$$
C_p=\frac1n(SSE+2p\hat\sigma^2).
$$

Thus, comparing to @eq-aic, $C_p$ and AIC are equivalent up to a constant scaling factor, and they rank models in essentially the same way.

From this perspective, $C_p$ can be viewed as the training error SSE plus a penalty term  $2p\hat\sigma^2$ that accounts for model complexity. More importantly, under the standard linear model assumptions, this adjustment corrects for the optimism of the training error, so that $C_p$ serves as an approximately unbiased estimator of the expected prediction error on new responses (at the same design points $X$). In this sense, $C_p$ estimates prediction error by adjusting the in-sample fit.





::: {.callout-caution}
## Number of parameters in $C_p$ and AIC

- In Mallows’ $C_p$, $p$ denotes the number of regression parameters, including the intercept.
- In AIC, the penalty uses the total number of estimated parameters. For a Gaussian linear model, this includes:
  - regression coefficients (including intercept), and
  - the error variance $\sigma^2$.

Thus, if $p$ counts regression parameters, AIC uses $k = p + 1$.

The difference between $C_p$ and AIC is only an additive constant across models. Therefore, when comparing models, these differences do not affect ranking, as long as a consistent definition is used.
:::



### PRESS
PRESS stands for **prediction sum of squares**. It measures how well a model predicts new observations by using leave-one-out prediction errors.

For each observation:

1. remove that observation,
2. fit the model using the remaining $n-1$ observations,
3. predict the omitted response,
4. compute the squared prediction error.

Then add all these squared prediction errors:

$$
PRESS = \sum_{i=1}^n (y_i-\hat y_{(i)})^2,
$$

where $\hat y_{(i)}$ is the predicted value for observation $i$ from the model fit without observation $i$.

Interpretation:

- A smaller PRESS means better out-of-sample predictive ability.
- Unlike ordinary SSE, PRESS directly focuses on prediction rather than only fit to the current sample.

::: {#exm-}

- Model A has $PRESS=85$
- Model B has $PRESS=62$

Then Model B is preferred for prediction, because its leave-one-out prediction errors are smaller overall.
:::



::: {#exm-}

We still use this generated dataset as an example.

```{r}
set.seed(123)
n <- 30
x1 <- runif(n, 1, 10)
x2 <- runif(n, 1, 10)
y <- 20 + 5 * x1 - 4 * x2 + rnorm(n, 0, 5)

dat <- data.frame(x1 = x1, x2 = x2, y = y)
```

```{r}
y_i <- numeric(n)
for (i in 1:n) {
    fit_i <- lm(y ~ x1 + x2, data = dat[-i, ])
    y_i[i] <- predict(fit_i, newdata = dat[i, ])
}
press_manual <- sum((y - y_i)^2)
press_manual
```

We could use `olsrr::ols_press` to compute PRESS. 

```{r}
#| warning: false
library(olsrr)
ols_press(fit)
```


:::


### Summary


So in practice:

- use $R^2$, adjusted $R^2$, or MSE to study fit,
- use AIC to balance fit and complexity,
- use $C_p$ to compare subset models,
- use PRESS when prediction is the main goal.

<!-- 
### BIC

The Bayesian Information Criterion (BIC) is very similar:

$$
BIC = -2\log L + k\log(n),
$$

where $n$ is the sample size. Since $\log(n)>2$ for most practical sample sizes, BIC usually penalizes extra variables more heavily than AIC.

Important points:

- AIC and BIC are used to compare models fit to the **same response on the same data**.
- Smaller values are preferred.
- These numbers are not meaningful by themselves; only **differences** between candidate models matter.

For a very easy example, suppose we compare two regression models fit to the same dataset:

- Model 1: $AIC=120$, $BIC=126$
- Model 2: $AIC=116$, $BIC=124$

Then Model 2 is preferred by both criteria because both numbers are smaller.

Now suppose instead we have:

- Model 1: $AIC=116$, $BIC=120$
- Model 2: $AIC=114$, $BIC=127$

Then AIC prefers Model 2, but BIC prefers Model 1. This usually means Model 2 fits a little better, but Model 1 is more parsimonious. In practice, AIC often leans toward prediction, while BIC often leans toward simpler models. -->






















## Stepwise Regression



Stepwise regression is a variable screening procedure that builds a model by repeatedly:

- **removing** predictors that are no longer useful, and
- **adding** useful predictors


At each step, the procedure examines statistical significance using a $t$-test (or the equivalent partial $F$-test) for individual regression coefficients.


A common stepwise procedure works like this:

1. Start with no predictors in the model.
2. Among the predictors in the model, compute test statistics and p-values.
3. Remove **any** variable whose p-value is too large, provided it is not significant enough to remove.
4. Among the predictors not yet in the model, compute test statistics and p-values.
5. Add **one** variable with the smallest p-value, provided it is significant enough to enter.
6. Repeat until no variable can be added or removed.

A variable should meet certain criteria to enter or remove. The most obvious one is that whether it is significant, or whether its p-value is smaller than $\alpha$ or not. However, in order to make the process more flexible, this threshold is modified to be 2 numbers.

- A variable can be removed if it is bigger than alpha-to-remove $\alpha_{r}$
- When increasing $\alpha_r$, variables are harder to be removed. 
- A variable can be added if it is smaller than alpha-to-enter $\alpha_{a}$
- When decreasing $\alpha_a$, variables are harder to be added. 
- In practice, one usually chooses $\alpha_a\leq \alpha_r$, often with values like $\alpha_a=0.05$ and $\alpha_r=0.10$.








<!-- 
If $\alpha_a=\alpha_r$, the procedure reduces to a simpler rule but becomes less flexible.

Stepwise methods are heuristic and may not yield the optimal model; they should be used cautiously (especially in inference settings). -->




<!-- So, unlike pure forward selection, stepwise regression allows a variable to enter first and then later be removed. -->

<!-- ### Stepwise regression steps in words

### Step 1

- No variable is currently in the model, so nothing can be removed.
- Test each predictor individually.
- Add the variable with the largest absolute $t$-statistic (equivalently, the smallest p-value), assuming it meets the entry criterion.

### Step 2

- Check whether the variable currently in the model should be removed.
- If not, examine the remaining predictors.
- Add the one that contributes the most, again based on the relevant test.

### Step 3 and beyond

- Re-check all variables currently in the model.
- Remove any variable whose p-value is too large.
- Then test the variables not yet in the model and add the most useful one.
- Continue until no predictor qualifies for entry or removal. -->

### Important notes about stepwise regression

Stepwise regression is often described as an **objective** screening method, but it has important limitations.

1. Many tests are performed: A very large number of hypothesis tests are carried out during the procedure. Because of this, the chance of making one or more Type I errors (including unimportant variables) or Type II errors (excluding important variables) can become quite high.
2. Usually only main effects are screened. Interaction terms and higher degree terms should be considered later. 
3. It does not search every possible subset. The search is a greedy algorithm that might get stuck in local optimum. 



::: {.callout-note}
The method covered here is the traditional p-value based stepwise regression. There are other versions that will be covered later.
:::




### Forward selection and backward elimination

Stepwise regression is closely related to two other variable screening strategies.

1. Forward selection: Forward selection starts with **no predictors** in the model.
   - At each step, add the most useful predictor.
   - Once a variable enters, it is **never removed**.
2. Backward elimination: Backward elimination starts with the **full model** containing all candidate predictors.
   - At each step, remove the least useful nonsignificant predictor.
   - Continue until all remaining predictors are statistically significant.
   - This method requires that the initial full model can actually be fit.

So stepwise selection combine both ideas:

- add predictors like forward selection,
- but after each addition, also check whether any existing predictor should be removed.

### R code

In R, we could use `olsrr::ols_step_both_p` to perform the stepwise regression.



::: {#exm-}


<details>
<summary><strong>Click to expand.</strong></summary>


The [dataset](EXAMPLE.csv) contains 10 predictors and 1 response variable.

```{r}
#| echo: false
#| eval: false

# set.seed(123)

# n <- 120

# # ---- independent vars ----
# X1 <- rnorm(n, 1, 2)
# X2 <- rnorm(n, 1, 1.5)
# X3 <- 0.6 * X1 + rnorm(n, 0, 0.1)
# X5 <- 0.7 * X2 + rnorm(n)
# X4 <- X1 - X2 + rnorm(n, 0, 0.5)
# X7 <- rbinom(n, 1, 0.5)
# X6 <- rnorm(n)
# X8 <- rnorm(n) + X6^2
# X9 <- rnorm(n) + X3 * X5 + X7
# X10 <- rnorm(n) - X2 * X4

# # ---- response ----
# eps <- rnorm(n, 0, 2)

# Y <- 3 + 2 * X1 - 1.8 * X2 - 1.1 * X6^2 + 1.1 * X8 - X10 + X9 + eps

# dat <- data.frame(
#     Y, X1, X2, X3, X4, X5, X6, X7, X8, X9, X10
# )
# fit <- lm(Y ~ ., data = dat)
# ols_step_both_p(fit, details = FALSE)
# write.csv(dat, "EXAMPLE.csv", row.names = FALSE)
```

```{r}
dat <- read.csv("EXAMPLE.csv")
head(dat)
```

We first use `lm` to get a 1st order model with all predictors.
```{r}
fit <- lm(Y ~ ., data = dat)
summary(fit)
```

There are 10 variables presented. We could use stepwise regression to screen variables.

```{r}
#| warning: false
library(olsrr)
ols_step_both_p(fit)
```
From this process, `X10 + X9 + X6 + X1 + X2 + X8` are our choice of variables.

</details>

:::

<!-- # Example: Executive Salary Data

A typical use of stepwise regression is when we have many candidate predictors and want to narrow the field before building a more refined model.

In the executive salary example described in the slides:

- the response variable is the **natural log of executive salary**,
- there are **10 potential predictors**,
- and the goal is to identify the most important main effects before considering a more complicated final model. -->

<!-- This is exactly the kind of situation where stepwise regression can serve as a first screening tool. -->

### Other versions

There are many issues about the traditional p-value based stepwise regression. Therefore we other other versions of the idea. `step` in base R is one of them.

There are two major changes here. 

1. The order of adding/removing variables are changed. During each step, `step` will compare the current model with each model that has one more varialbe, and with each model that has one few variable, and choose the best one for the next iteration. In other words, it perform adding and removing simutanously.
2. The best in this function is the smallest AIC. 


::: {#exm-}

<details>
<summary><strong>Click to expand.</strong></summary>

```{r}
step(fit, direction = "both")
```

If we want to start from `y~x1+x2+x3+x4` and to search until `x10`, we need to first fit a model with 4 variables, and define the upper bound to be 10 variables.


```{r}
fit_sub <- lm(Y~X1+X2+X3+X4, data=dat)
step(
    fit_sub,
    scope = list(
        lower = ~1,
        upper = ~ X1+X2+X3+X4+X5+X6+X7+X8+X9+X10
    )
)
```


</details>

:::















## All-Possible-Regressions Selection Procedure


Forward selection, backward elimination, and stepwise regression all build models one variable at a time. None of them guarantees that the best model with a given number of predictors will be found.

Aall-possible-regressions selection, or alternatively named best-subsets regression, is to consider all possible regression models formed from the candidate predictors. The idea is very simple: for each possible model size, search all possible models and identify the best subset according to a chosen criterion by brute force.

Typical criteria contain

- $R^2$ or Adjusted $R^2$
- MSE
- Mallows' $C_p$ or AIC
- PRESS 

In R, we could use `leaps::regsubsets` to search for the variables.


::: {#exm-}

<details>
<summary><strong>Click to expand.</strong></summary>

```{r}
#| warning: false

library(leaps)
dat <- read.csv("EXAMPLE.csv")
```


```{r}
models <- regsubsets(Y ~ ., data = dat, nvmax = 10)
models_summary <- summary(models)
```

We could then check the which subset is the best according to different creteria.


```{r}
best_cp <- which.min(models_summary$cp)
plot(models_summary$cp, xlab = "Number of varaiables", type = "l")
points(best_cp, models_summary$cp[best_cp], col = "red", cex = 2, pch = 20)
```



```{r}
best_adjr2 <- which.max(models_summary$adjr2)
plot(models_summary$adjr2, xlab = "Number of varaiables", type = "l")
points(best_adjr2, models_summary$adjr2[best_adjr2], col = "red", cex = 2, pch = 20)
```

```{r}
best_rsq <- which.max(models_summary$rsq)
plot(models_summary$rsq, xlab = "Number of varaiables", type = "l")
points(best_rsq, models_summary$rsq[best_rsq], col = "red", cex = 2, pch = 20)
```

```{r}
best_rss <- which.min(models_summary$rss)
plot(models_summary$rss, xlab = "Number of varaiables", type = "l")
points(best_rss, models_summary$rss[best_rss], col = "red", cex = 2, pch = 20)
```


</details>

:::








::: {.callout-note collapse="true"}
## Stepwise Regression in Statistical Learning

This description follows @Jam2021. The key idea is to approximate best subset selection using a greedy procedure.

1. For each model size (number of predictors), stepwise procedures generate a set of candidate models by adding or removing variables.
2. Among models of the same size, we select the “best” model based on training fit (e.g., largest $R^2$ or smallest SSE). This yields one representative model for each model size.
3. We then compare these models across different sizes using an estimate of test error, such as AIC, $C_p$, or really using a validation set to estimate the test error. 

Note that this stepwise regression is actually a variation of the best-of-subset method. Therefore its R implementation is done through `leaps::regsubsets`.


```{r}
#| warning: false

library(leaps)
fit_fwd <- regsubsets(Y ~ ., data = dat, nvmax = 10, method = "forward")
summary(fit_fwd)
```


```{r}
fwd_sum <- summary(fit_fwd)
best_cp <- which.min(fwd_sum$cp)
best_cp
```

```{r}
plot(fwd_sum$cp, xlab='Number of varaiables', type='l')
points(best_cp, fwd_sum$cp[best_cp], col = "red", cex = 2, pch = 20)
```

Note that `regsubsets` doesn't use AIC since AIC and $C_p$ are essentially the same in the case of linear regression. 

:::









## Caveats

Both stepwise regression and best-subsets regression should be used carefully.

- Test error rates can be high
- These are screening tools, not full model-building procedures, since usually, analysts do not include interaction terms and higher-order terms in the initial candidate list.
- They may produce nonsensical models, therefore common sense still matters. 

You should not abandon:

- subject-matter knowledge,
- common sense,
- intuition,
- and interpretability.

A statistically selected model is not automatically the right model. 

A good way to use these methods is:

1. Start with a long list of possible predictors.
2. Use stepwise regression or best-subsets regression to screen for the most important variables.
3. Then begin the real model-building phase:
   - check assumptions,
   - examine residual plots,
   - consider transformations,
   - add interactions or higher-order terms when scientifically justified,
   - and compare models thoughtfully.

So these methods are best viewed as starting points, not final answers.



::: {.callout-tip}
## LASSO
LASSO is fundamentally different from ordinary least squares (OLS). While OLS focuses on unbiased estimation and supports classical inference (such as confidence intervals and hypothesis tests), LASSO introduces a penalty that shrinks coefficients and can set some of them exactly to zero. This makes LASSO particularly effective for prediction and variable selection, especially in settings with many or highly correlated predictors. 

However, because the estimates are biased and the selection process is data-dependent, standard inferential tools (like p-values and standard errors) are no longer valid in the usual sense. For this reason, LASSO is not emphasized in introductory regression courses. Instead, it plays a central role in modern statistical learning, and we will study it in more detail in other courses which focused on prediction and model regularization if there is an opportunity.
:::