



# Variable selection


{{< include ../math.qmd >}} 

When we build a multiple regression model, we often begin with many possible predictors. The main question is: **which variables should be included in the model?**

This chapter introduces two systematic methods for screening variables:

1. **Stepwise regression**
2. **All-possible-regressions selection** (also called **best-subsets regression**)

These methods are intended to help identify the most important predictors before the more careful model-building phase begins.

<!-- ## Why Use a Variable Screening Method

In many applications, researchers collect a response variable $y$ together with a large number of potential independent variables $x_1, x_2, \dots, x_p$. Examples include predicting:

- firm profit,
- college GPA,
- inflation rate,
- salary,
- or other outcomes of interest.

When there are many candidate predictors, it can be difficult to decide which variables should enter the regression model. Variable screening methods provide a more systematic way to reduce the list.

The two methods emphasized in this chapter are:

- **Stepwise regression**
- **All-possible-regressions selection** -->

## Model Selection Criteria
Variable selections can be treated as a model comparison process. Therefore we need to discuss criteria on comparing models.

### What we already covered
1. p-value: when comparing nested models, the p-value can be used to determine whether the added variables are significant as a group.
2. $R^2$ and $R_a^2$: the proportion of the varaiance that can be explained. This is one of the metrics that is used to determine the performance of a model.

When talking about approximation ability, some tend to use MSE. This is very similar to $R^2$ and $R_a^2$ since they are computed using MSE.

### AIC
The Akaike Information Criterion (AIC) is a statistical metric used to evaluate the performance of a model. It balances model fit (goodness of fit) with complexity to avoid over or underfitting. Lower AIC values indicate a better, more efficient model.

A common formula is

$$
AIC = -2 L + 2k,
$$

where $L$ is the likelihood of the fitted model and $k$ is the number of estimated parameters. The first term rewards good fit, while the second term  penalizes model complexity. In the case of linear regression, AIC is 
$$
AIC=n\log\frac{SSE}{n}+2k
$$
where $SSE$ is the sum of squared errors of the model, $n$ is the number of observations and $k$ is the number of estimated predictors.



::: {.callout-note collapse="true"}
## The rough story of AIC

AIC is derived as an estimate of expected out-of-sample prediction error, measured via Kullback–Leibler (KL) divergence.

1. Let $g(y)$ be the true but unknown data-generating distribution coming from the real world, and $f(y\mid\theta)$ be the candidate model to estimate $g(y)$, where $\theta$ is the parameters vector of the model. 
2. Akaike's idea to measure the difference between $g$ and $f$ is KL divergence $D_{KL}$. We would like to find the parameters $\hat\theta$ such that $D_{KL}(g\,\|\,f_{\theta})$ is minimized.
$$
\begin{split}
D_{KL}(g\,\|\,f)&=\int g(y)\log\frac{g(y)}{f(y\mid \theta)}\dl3y=\underbrace{\int g(y)\log g(y)\dl3y}_{\text{constant}}-\int g(y)\log f(y\mid\theta)\dl3y\\
&=\int g(y)\log g(y)\dl3y-\Exp_g[\log f(Y\mid\theta)].
\end{split}
$$
So we need to find the $\hat\theta$ such that $\Exp_g[\log f(Y\mid\theta)]$ is maximized.
3. Since we don't know the real $g$, but we have samples from real world $Y_1,\ldots,Y_n$, so we maximize the sample log-likelihood:
$$
\ell(\theta)=\sum_{i=1}^n\log f(Y_i\mid\theta).
$$
This gives the MLE estimator $\hat\theta$.
4. However $\ell(\hat\theta)$ is optimistically biased for the out-of-sample log-likelihood $\ell^*(\hat\theta)$, because the same data is used for fitting and evaluation.
5. Akaike’s key result (asymptotic bias correction) proved that under some regularity conditions, 
$$
\Exp_g[\ell(\hat\theta)]\approx \Exp_g[\ell^*(\hat\theta)]+k,
$$
where $k$ is the number of estimated parameters. Therefore 
$$
\Exp_g[\ell^*(\hat\theta)]\approx \Exp_g[\ell(\hat\theta)-k].
$$
So a natural estimator for $\ell^*(\hat\theta)$ is $\ell(\hat\theta)-k$. 
6. The estimator suggests the definition
$$
AIC=-2\ell(\hat\theta)+2k.
$$
The factor $-2$ is conventional. It puts AIC on the deviance scale, making it consistent with likelihood-ratio theory and convenient for model comparison.
7. Consider a MLR model, $Y=X^T\beta+\varepsilon$, where $\varepsilon\sim N(0,\sigma^2)$. Then based on the MLE estimator calculation, the log-likelihood is
$$
\ell(\beta,\sigma^2)= -\frac{n}{2}\log(2\pi)- \frac{n}{2}\log(\sigma^2)- \frac{1}{2\sigma^2}\mathrm{SSE}.
$$
Using the MLE $\hat{\sigma}^2 = \mathrm{SSE}/n$, we obtain
$$
\ell(\hat{\theta})= -\frac{n}{2}\left[\log(2\pi) + 1 + \log\left(\frac{\mathrm{SSE}}{n}\right)\right]=-\frac{n}{2}\log\left(\frac{\mathrm{SSE}}{n}\right)- \underbrace{\frac{n}{2}\left(1 + \log(2\pi)\right)}_{\text{irrelevant constant}}.
$$
Since the constant does not depend on the model, for comparing linear regression models we often use
$$
\ell(\hat{\theta}) \propto -\frac{n}{2}\log\left(\frac{\mathrm{SSE}}{n}\right).
$$
8. Multiplying by $-2$ gives the final form of AIC in MLR:
$$
AIC=n\log\qty(\frac{SSE}n)+2k.
$$





:::






The Bayesian Information Criterion (BIC) is very similar:

$$
BIC = -2\log L + k\log(n),
$$

where $n$ is the sample size. Since $\log(n)>2$ for most practical sample sizes, BIC usually penalizes extra variables more heavily than AIC.

Important points:

- AIC and BIC are used to compare models fit to the **same response on the same data**.
- Smaller values are preferred.
- These numbers are not meaningful by themselves; only **differences** between candidate models matter.

For a very easy example, suppose we compare two regression models fit to the same dataset:

- Model 1: $AIC=120$, $BIC=126$
- Model 2: $AIC=116$, $BIC=124$

Then Model 2 is preferred by both criteria because both numbers are smaller.

Now suppose instead we have:

- Model 1: $AIC=116$, $BIC=120$
- Model 2: $AIC=114$, $BIC=127$

Then AIC prefers Model 2, but BIC prefers Model 1. This usually means Model 2 fits a little better, but Model 1 is more parsimonious. In practice, AIC often leans toward prediction, while BIC often leans toward simpler models.





### C_p
Mallows' $C_p$ is another criterion for comparing subset regression models. It tries to balance lack of fit and model size.

For a candidate subset model with $p$ parameters (including the intercept), the statistic is

$$
C_p=\frac{SSE_p}{MSE_{\text{full}}}-(n-2p),
$$

where:

- $SSE_p$ is the error sum of squares for the candidate subset model,
- $MSE_{\text{full}}$ is the mean square error from the full model,
- $n$ is the sample size.

How to interpret it:

- A **smaller** $C_p$ is better.
- A good model often has $C_p$ **close to $p$**.
- If $C_p$ is much larger than $p$, the subset model may have substantial bias because important variables are missing.

Easy example:

- Model A has $p=3$ parameters and $C_p=12$
- Model B has $p=5$ parameters and $C_p=5.4$

Model B is usually preferred. Its $C_p$ is small and very close to $p=5$, which suggests that the model is capturing the important predictors without too much bias. Model A has a much larger $C_p$ than $p=3$, so it likely leaves out useful variables.


### PRESS
PRESS stands for **prediction sum of squares**. It measures how well a model predicts new observations by using leave-one-out prediction errors.

For each observation:

1. remove that observation,
2. fit the model using the remaining $n-1$ observations,
3. predict the omitted response,
4. compute the squared prediction error.

Then add all these squared prediction errors:

$$
PRESS = \sum_{i=1}^n (y_i-\hat y_{(i)})^2,
$$

where $\hat y_{(i)}$ is the predicted value for observation $i$ from the model fit without observation $i$.

Interpretation:

- A smaller PRESS means better out-of-sample predictive ability.
- Unlike ordinary SSE, PRESS directly focuses on prediction rather than only fit to the current sample.

Easy example:

- Model A has $PRESS=85$
- Model B has $PRESS=62$

Then Model B is preferred for prediction, because its leave-one-out prediction errors are smaller overall.

So in practice:

- use $R^2$, adjusted $R^2$, or MSE to study fit,
- use AIC/BIC to balance fit and complexity,
- use $C_p$ to compare subset models,
- use PRESS when prediction is the main goal.




## Stepwise Regression



Stepwise regression is a variable screening procedure that builds a model by repeatedly:

- **removing** predictors that are no longer useful, and
- **adding** useful predictors


At each step, the procedure examines statistical significance using a $t$-test (or the equivalent partial $F$-test) for individual regression coefficients.


A common stepwise procedure works like this:

1. Start with no predictors in the model.
2. Among the predictors in the model, compute test statistics and p-values.
3. Remove **any** variable whose p-value is too large, provided it is not significant enough to remove.
4. Among the predictors not yet in the model, compute test statistics and p-values.
5. Add **one** variable with the smallest p-value, provided it is significant enough to enter.
6. Repeat until no variable can be added or removed.

A variable should meet certain criteria to enter or remove. The most obvious one is that whether it is significant, or whether its p-value is smaller than $\alpha$ or not. However, in order to make the process more flexible, this threshold is modified to be 2 numbers.

- A variable can be removed if it is bigger than alpha-to-remove $\alpha_{r}$
- When increasing $\alpha_r$, variables are harder to be removed. 
- A variable can be added if it is smaller than alpha-to-enter $\alpha_{a}$
- When decreasing $\alpha_a$, variables are harder to be added. 
- In practice, one usually chooses $\alpha_a\leq \alpha_r$, often with values like $\alpha_a=0.05$ and $\alpha_r=0.10$.








<!-- 
If $\alpha_a=\alpha_r$, the procedure reduces to a simpler rule but becomes less flexible.

Stepwise methods are heuristic and may not yield the optimal model; they should be used cautiously (especially in inference settings). -->




<!-- So, unlike pure forward selection, stepwise regression allows a variable to enter first and then later be removed. -->

<!-- ### Stepwise regression steps in words

### Step 1

- No variable is currently in the model, so nothing can be removed.
- Test each predictor individually.
- Add the variable with the largest absolute $t$-statistic (equivalently, the smallest p-value), assuming it meets the entry criterion.

### Step 2

- Check whether the variable currently in the model should be removed.
- If not, examine the remaining predictors.
- Add the one that contributes the most, again based on the relevant test.

### Step 3 and beyond

- Re-check all variables currently in the model.
- Remove any variable whose p-value is too large.
- Then test the variables not yet in the model and add the most useful one.
- Continue until no predictor qualifies for entry or removal. -->

### Important notes about stepwise regression

Stepwise regression is often described as an **objective** screening method, but it has important limitations.

1. Many tests are performed: A very large number of hypothesis tests are carried out during the procedure. Because of this, the chance of making one or more Type I errors (including unimportant variables) or Type II errors (excluding important variables) can become quite high.
2. Usually only main effects are screened. Interaction terms and higher degree terms should be considered later. 
3. It does not search every possible subset. The search is a greedy algorithm that might get stuck in local optimum. 


### Forward selection and backward elimination

Stepwise regression is closely related to two other variable screening strategies.

1. Forward selection: Forward selection starts with **no predictors** in the model.
   - At each step, add the most useful predictor.
   - Once a variable enters, it is **never removed**.
2. Backward elimination: Backward elimination starts with the **full model** containing all candidate predictors.
   - At each step, remove the least useful nonsignificant predictor.
   - Continue until all remaining predictors are statistically significant.
   - This method requires that the initial full model can actually be fit.

So stepwise selection combine both ideas:

- add predictors like forward selection,
- but after each addition, also check whether any existing predictor should be removed.

### R code

In R, we could use `olsrr::ols_step_both_p` to perform the stepwise regression.



::: {#exm-}

The [dataset](EXAMPLE.csv) contains 10 predictors and 1 response variable.

```{r}
#| echo: false
#| eval: false

# set.seed(123)

# n <- 120

# # ---- independent vars ----
# X1 <- rnorm(n, 1, 2)
# X2 <- rnorm(n, 1, 1.5)
# X3 <- 0.6 * X1 + rnorm(n, 0, 0.1)
# X5 <- 0.7 * X2 + rnorm(n)
# X4 <- X1 - X2 + rnorm(n, 0, 0.5)
# X7 <- rbinom(n, 1, 0.5)
# X6 <- rnorm(n)
# X8 <- rnorm(n) + X6^2
# X9 <- rnorm(n) + X3 * X5 + X7
# X10 <- rnorm(n) - X2 * X4

# # ---- response ----
# eps <- rnorm(n, 0, 2)

# Y <- 3 + 2 * X1 - 1.8 * X2 - 1.1 * X6^2 + 1.1 * X8 - X10 + X9 + eps

# dat <- data.frame(
#     Y, X1, X2, X3, X4, X5, X6, X7, X8, X9, X10
# )
# fit <- lm(Y ~ ., data = dat)
# ols_step_both_p(fit, details = FALSE)
# write.csv(dat, "EXAMPLE.csv", row.names = FALSE)
```

```{r}
dat <- read.csv("EXAMPLE.csv")
head(dat)
```

We first use `lm` to get a 1st order model with all predictors.
```{r}
fit <- lm(Y ~ ., data = dat)
summary(fit)
```

There are 10 variables presented. We could use stepwise regression to screen variables.

```{r}
#| warning: false
library(olsrr)
ols_step_both_p(fit)
```
From this process, `X10 + X9 + X6 + X1 + X2 + X8` are our choice of variables.

:::

<!-- # Example: Executive Salary Data

A typical use of stepwise regression is when we have many candidate predictors and want to narrow the field before building a more refined model.

In the executive salary example described in the slides:

- the response variable is the **natural log of executive salary**,
- there are **10 potential predictors**,
- and the goal is to identify the most important main effects before considering a more complicated final model. -->

<!-- This is exactly the kind of situation where stepwise regression can serve as a first screening tool. -->


### Other metrics





```{r}
library(olsrr)
ols_step_both_p(fit, details = TRUE)
```


```{r}
step(fit)
```




```{r}
library(leaps)
models <- regsubsets(Y ~ ., data = dat, nvmax = 10)
summary(models)
```

```{r}
plot(summary(models)$adjr2)
```


# 6.3 All-Possible-Regressions Selection Procedure

## Basic idea

Forward selection, backward elimination, and stepwise regression all build models **one variable at a time**. None of them guarantees that the best model with a given number of predictors will be found.

An alternative is to consider **all possible regression models** formed from the candidate predictors.

This approach is called:

- **all-possible-regressions selection**, or
- **best-subsets regression**.

Its main advantage is that, for each possible model size, it can identify the best subset according to a chosen criterion.

## Why use best-subsets regression?

Suppose there are $p$ candidate predictors.

Then for:

- 1 predictor,
- 2 predictors,
- 3 predictors,
- and so on,

we can compare all models of that size and select the best one according to some measure of quality.

This gives more complete information than a purely sequential method.

# Criteria Used in Best-Subsets Regression

Several criteria may be used to compare subset models.

## $R^2$ criterion

A larger $R^2$ indicates a better fit to the observed data.

However, $R^2$ always increases when more predictors are added, so it tends to favor larger models.

## Adjusted $R^2$ criterion

Adjusted $R^2$ penalizes model size.

In general:

- larger adjusted $R^2$ is better.

Because adjusted $R^2$ increases only when the mean square error decreases enough to justify the added complexity, it is often more useful than ordinary $R^2$.

## MSE criterion

A smaller **mean square error (MSE)** is preferred.

Searching for a model with minimum or near-minimum MSE is equivalent to searching for a model with high adjusted $R^2$.

## Mallows' $C_p$ criterion

A good subset model should have:

- a **small** value of $C_p$, and
- a value of $C_p$ **near** the number of parameters in the model.

A $C_p$ value near the number of parameters suggests that the subset model has little bias.

## PRESS criterion

PRESS stands for **prediction sum of squares**.

It is based on leave-one-out prediction errors:

$$
PRESS = \sum_{i=1}^n (y_i - \hat y_{(i)})^2,
$$

where $\hat y_{(i)}$ is the predicted value for observation $i$ obtained from a model fit **without** using observation $i$.

A smaller PRESS indicates better predictive performance.

# Example: Best-Subsets for Executive Salary

In the executive salary example, best-subsets regression is used to examine the 10 potential predictors more thoroughly.

The goal is again to identify the most important predictors for the natural log of salary, but now the procedure compares possible subsets more directly rather than following only a sequential path.

Plots and software output for $R^2$, adjusted $R^2$, $C_p$, and related criteria can help identify a promising subset size and a promising combination of variables.

# 6.4 Caveats

Both stepwise regression and best-subsets regression should be used carefully.

## Main caveats

### 1. Error rates can be high

Because many models and many tests are examined, the probability of making at least one Type I or Type II error can be quite large.

### 2. These are screening tools, not full model-building procedures

Usually, analysts do **not** include interaction terms and higher-order terms in the initial candidate list.

As a result, these procedures often do **not** perform the full intellectual work of model building.

### 3. They may produce nonsensical models

Even if some interaction or higher-order terms are included, automated methods can still select a model that makes little scientific or practical sense.

### 4. Common sense still matters

You should not abandon:

- subject-matter knowledge,
- common sense,
- intuition,
- and interpretability.

A statistically selected model is not automatically the right model.

# Practical Advice

A good way to use these methods is:

1. Start with a long list of possible predictors.
2. Use **stepwise regression** or **best-subsets regression** to screen for the most important variables.
3. Then begin the **real model-building phase**:
   - check assumptions,
   - examine residual plots,
   - consider transformations,
   - add interactions or higher-order terms when scientifically justified,
   - and compare models thoughtfully.

So these methods are best viewed as **starting points**, not final answers.

# Quick Comparison Table

| Method | How it starts | How it proceeds | Strength | Limitation |
|---|---|---|---|---|
| Forward selection | No predictors | Adds variables only | Simple | Cannot remove variables once added |
| Backward elimination | Full model | Removes variables only | Starts from full information | Requires fitting the full model |
| Stepwise selection | No predictors | Adds and removes | More flexible than forward selection | Still does not check all subsets |
| Best-subsets regression | Considers all subset sizes | Compares all candidate models | Can find best subset for each size | Computationally heavier; may still ignore scientific meaning |

# R Notes

Below are common R approaches for these methods.

## Stepwise regression

```r
fit_full <- lm(y ~ x1 + x2 + x3 + x4, data = dat)
fit_null <- lm(y ~ 1, data = dat)

step(fit_null,
     scope = list(lower = fit_null, upper = fit_full),
     direction = "both")
```

## Forward selection

```r
step(fit_null,
     scope = list(lower = fit_null, upper = fit_full),
     direction = "forward")
```

## Backward elimination

```r
step(fit_full, direction = "backward")
```

## Best-subsets regression

```r
library(leaps)

best <- regsubsets(y ~ x1 + x2 + x3 + x4, data = dat, nvmax = 4)
summary(best)
```

# Final Takeaway

Variable screening methods are useful when there are many possible predictors, but they should be used with caution.

- **Stepwise regression** is a convenient automatic screening procedure.
- **Best-subsets regression** is more exhaustive and can identify the best model for a given size.
- Neither method should replace thoughtful model building.

The best practice is to use these procedures to **screen variables first**, then build the final regression model using statistical reasoning, subject knowledge, and diagnostic checking.
