---
title: "Homework 1"
# author: "Type Your Name Here"
date: today
format: pdf
---


## Question 1. Predicting home sales price (HOMESALES.csv)

(a) Write down the straight_line model.




(b) 
- Draw the scatter plot of Sale_Price vs Market_val in R. 

```{r}

### Read the data set in .csv


### Plot the scatter plot

```

- Does it appear that a straight-line model will be an appropriate fit to the data?



(c) 

-Fit the regression model in R. 

```{r}
### Fit the model and find the summary and anova table

```


- Write down the equation of the best-fitting line through the data.





(d) Interpret the $y$-intercept of the least squares line. Does it hava a practical meaning for this application? Explain.





(e) 

- Interpret the slope of the least squares line.





- Over what range of $x$ is the interpretation meaningful?



(f) Use the least squares model fitted to estimate the mean sale price of a property appraised at \$300,000.





## Question 2. Game performance of water polo players (POLO.csv)

(a) Find the equation of the least squares line.

```{r}




```

- The estimated regression equation is:


(b) Give a practical interpretation (if possible) of the $y$-intercept of the line.





(c) Give a practical interpretaiton (if possible) of the slope of the line.






## Question 3. Spreading rate of spilled liquid (LIQUIDSPILL.csv)

```{r}
### Read the data set in .csv and draw scatter plot

```


- Do the data indicate that the mass of the spill tends to diminish as time increase?



```{r}
### Fit the simple linear regression model

```

- If the data indicate the trend, how much will the mass diminish each minutes?







## Question 4. Sweetness of orange juice (OJUICE.csv)

(a) Fit the model and find the values of SSE, $s^2$, and $s$ for this regression.

```{r}

```


- SSE =
- $s^2$ =
- $s$ =


(b) Estimate $\sigma^2$, the variance of the random error term in the model.



(c) Estimate $\sigma$, the standard deviation of the random error term in the model. 


(d) Explain why it is difficult to give a practical interpretation to $s^2$, the estimate of $\sigma^2$.



(e) Give a practical interpretation of the value of $s$. 

