---
title: "Homework 3"
author: "Type Your Name Here"
date: "10/1/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, warning = FALSE)
```


## Question 1. Reality TV and cosmetic surgery (Data set: BDYIMG))

(a) Fit the first-order model, $E(y)=\beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_3+\beta_4x_4$, to the data in the file. Give the least squares prediction equation

```{r}

### Read the data



### Fit the MLR model



```

Answer: The least squares prediction equation is



(b) Interpret the $\beta$ estimates in the words of the problem. (Yes, you need to find the values of $\hat\beta_0, \hat\beta_1, \hat\beta_2, \hat\beta_3, \hat\beta_4$ and interpret them one by one.)


Answer:



(c) Is the overall model statistically useful for predicting desire to have cosmetic surgery? Test using $\alpha=.01$. (When performing the hypothesis testing, do not forget to write down the hypotheses first!)

Answer:





## Question 2. Arsenic in groundwater (Data set: ASWELLS)

```{r}
### Import data and fit the MLR model




```


(a) Write a first-order model for arsenic level ($y$) as a function of latitude, longitude, and depth.

Answer:



(b) Fit the model to the data using the method of least squares.

Answer:




(c) Find the value of $\hat\beta_2$ and give a practical interpretation of $\hat\beta_2$.


Answer:



(d) Find the model standard deviation, $s$, and interpret its value.

Answer:



(f) Conduct a test of overall model utility at $\alpha=.05$. (When performing the hypothesis testing, do not forget to write down the hypotheses first!)

Answer:











