---
title: "HW 1"
# author: "Type Your Name Here"
# date: today
format: 
    pdf:
        include-in-header:
        - text: |
            \AtBeginDocument{
                \renewcommand{\theexercise}{\arabic{exercise}}
            }
params:
    show_solution: false
filters:
    - ../../solution-filter.lua
crossref:
    chapters: false
---



# Exercises


::: {#exr-}
## Learning the mechanics. 
Use the method of least squares to fit a straight line to these six data points:

| x | 1 | 2 | 3 | 4 | 5 | 6 |
|---|---|---|---|---|---|---|
| y | 2 | 4 | 5 | 4 | 2 | 7 |


(a) What are the least squares estimates of $\beta_0$ and $\beta_1$?
(b) Plot the data points and graph the least squares line on the scatterplot.
:::


::: {.answer}
(a)
- With hand:
  - $\bar x=3.5$, $\bar y=4$.
  
   | $x-\bar x$ | -2.5 | -1.5 | -0.5 | 0.5 | 1.5 | 2.5 |
    |---|---|---|---|---|---|---|
    | $y-\bar y$ | -2 | 0 | 1 | 0 | -2 | 3 |
  - $S_{xy}=\sum (x_i-\bar x)(y_i-\bar y)=9$, $S_{xx}=\sum (x_i-\bar x)^2=17.5$.
  - $\hat\beta_1=9/17.5\approx 0.5143$, $\hat\beta_0=\bar y-\beta_1\bar x=4-9/17.5\times3.5=2.2$.
- With R
```{r}
x <- c(1, 2, 3, 4, 5, 6)
y <- c(2, 4, 5, 4, 2, 7)
fit <- lm(y~x)
summary(fit)
```

(b) Ploting with R code.

```{r}
plot(x, y)
abline(coef = fit$coefficients, col='red')
```

:::





::: {#exr-}
## Predicting home sales price. 
Real estate investors, homebuyers, and homeowners often use the appraised (or market) value of a property as a basis for predicting sale price. Please look at the provided dataset [`MARKET.csv`](./datasets/market.csv). The first five and last five observations of the data set are listed in the accompanying table.

(a) Propose a simple linear model to relate the appraised market value $x$ to the sale price $y$.
(b) A scatterplot of the data is shown below. Does it appear that a straight-line model will be an appropriate fit to the data?
(c) A R simple linear regression printout is also shown below. Find the equation of the best-fitting line through the data on the printout.
(d) Interpret the $y$-intercept of the least squares line. Does it have a practical meaning for this application? Explain.
(e) Interpret the slope of the least squares line. Over what range of $x$ is the interpretation meaningful?
(f) Use the least squares model to estimate the mean sale price of a property appraised at $300,000.

```{r}
#| echo: false

df <- read.csv("datasets/MARKET.csv")
attach(df)
plot(market_value, sale_price)
```
```{r}
#| echo: false

fit <- lm(sale_price~market_value)
summary(fit)
```
:::


::: {#exr-}
A study shows that during a certain sport the mean heart rate $y$ and the maximal oxygen uptake $x$ might have relations. The dataset `SPORTHR.csv` shows $y$ (expressed as a percentage of maximum heart rate) and $x$
($V0_2$max). The data are shown in the table.


```{r}
#| echo: false

df <- read.csv("datasets/SPORTHR.csv")
print(df, row.names=FALSE)
```


(a) Find the equation of the least squares line.
(b) Give a practical interpretation (if possible) of the $y$-intercept of the line.
(c) Give a practical interpretation (if possible) of the slope of the line.
:::







::: {#exr-}
## Spreading rate of spilled liquid. 
A researcher studied the rate at which a spilled liquid will spread across a surface. The mass (in pounds) of the spill after a period of time ranging from 0 to 60 minutes is recorded and shown below (based on the dataset `SPILLS.csv`). Do the data indicate that the mass of the spill tends to diminish as time increases? If so, how much will the mass diminish each minute?

```{r}
#| echo: false

df <- read.csv("datasets/SPILLS.csv")
print(df, row.names=FALSE)
```
:::








::: {#exr-}
## Sweetness of orange juice. 
To study the sweetness of orange juices, researchers collect some data on the sweetness index ($y$) and the amount of pectin ($x$) in the orange juice (in $g/L$). The dataset is `ORANGEJUICE.csv`.


```{r}
#| echo: false

df <- read.csv("datasets/ORANGEJUICE.csv")
print(df, row.names=FALSE)
```



(a) Find the values of $SSE$, $s^2$, and $s$ for this regression.
(b) Estimate $\sigma^2$, the variance of the random error term in the model.
(c) Estimate $\sigma$, the standard deviation of the random error term in the model.
(d) Explain why it is difficult to give a practical interpretation to $s^2$, the estimate of $\sigma^2$.
(e) Give a practical interpretation of the value of $s$.
:::