---
title: "HW 2"
# author: "Type Your Name Here"
# date: today
format: 
    pdf:        
        pdf-engine: lualatex
        template-partials:
            - ../../doc-class.tex
        include-in-header:
            - text: |
                \usepackage{xcolor}
                \AtBeginDocument{
                    \renewcommand{\theexercise}{\arabic{exercise}}
                }
params:
    show_solution: false
filters:
    - ../../solution-filter.lua
crossref:
    chapters: false
---


::: {#exr-juice}
## Sweetness of orange juice. 
To study the sweetness of orange juices, researchers collect some data on the sweetness index ($y$) and the amount of pectin ($x$) in the orange juice (in $g/L$). The dataset is `ORANGEJUICE.csv`.


```{r}
#| echo: false
#| output: false

df <- read.csv("datasets/ORANGEJUICE.csv")
knitr::kable(df, format = "pipe")
```


(a) Fit the model and find a 90% confidence interval for the true slope of the line. Interpret the result.
(b) Fit the model and determine whether there is a positive or negative linear relationship between the amount of pectin $x$ and the sweetness $y$. That is, determine if there is sufficient evidence (at $\alpha=0.05$) to indicate that $\beta_1$, the slope of the straight-line model, is significantly different from zero. 

:::



::: {#exr-}
## Car program

A company bought a car or two each year. The data is shown in the file `COMPANYCAR.csv`.

(a) Fit the simple linear regression model, $E(y)=\beta_0+\beta_1$, to the data.
(b) List assumptions required for the regression analysis.
(c) Find the value of SSE.
(d) Find the estimated standard error of the regression model, $s$.
(e) Give a practical interpretation of $s$.
(f) Find a 95% confidence interval for the true slope of the line. 
(g) Interpret the confidence interval in (f).
(h) Find the $p$-value for testing $H_0:\beta_1 = 0$ versus $H_a:\beta_1\neq 0$. Use this result to test the simple linear regression model is statistically useful for predicting the annual cost using the year of initial operation. (Test using $\alpha=0.05$)
(i) Find and interpret the coefficient of determination, $R^2$.
(j) A researcher wants to estimate of the average annual cost of company cars with the year in 2020. Which interval is desired by the researcher, a 95% prediction interval for y or a 95% confidence interval for $E(y)$? Use R to calculate the desired interval.
(k) Give a practical interpretation of the interval in part (j).
:::



::: {#exr-}
## Fill in the blanks in the table and answer questions
Look at the output of the linear regression model. Fill in the blanks in the table and answer questions:


```{r}
#| echo: false
#| output: false
source("../../../anova_alt.R")
df <- read.csv("datasets/DATA.csv")
fit <- lm(y ~ x, data = df)
s <- capture.output(summary(fit))
s <- paste(s, collapse = "\n")
a <- capture.output(anova_alt(fit))
```


```{r}
#| echo: false
#| results: asis

s <- gsub(
    "12.295   <2e-16 \\*\\*\\*",
    "1. ____   2. ____",
    s
)

s <- gsub(
    "Residual standard error:\\s*[0-9\\.]+",
    "Residual standard error:  3. _______",
    s
)

a <- gsub(
    "29582640 29582640",
    "29582640  4. _______",
    a
)

a <- gsub(
    " 195710",
    "5. _______",
    a
)

cat("\\begin{verbatim}\n")
cat(paste(s, collapse = "\n"))
cat("\n\\end{verbatim}\n")

cat("\\begin{verbatim}\n")
cat(paste(a, collapse = "\n"))
cat("\n\\end{verbatim}\n")
```


(a) Please fill in the blanks in the table.
    (1) 
    (2)
    (3)
    (4)
    (5)
(b) Find and interpret the coefficient of determination, $r^2$.
(c) Calculate the coefficient of correlation, $r$.

:::



::: {#exr-}
## Position effects in memories

Here is an experiment. There are 9 words on screen and participants are requested to recall these words as many as possible. The results are stored in the file `WORDMEMORY.csv`: each row represens a word and the recall rate of the word at this position is recorded.

(a) Find a 99% confidence interval for the mean recall proportion for words in the fifth position. Interpret the result.
(b) Find a 99% prediction interval for the recall proportion of a particular word in the fifth position. Interpret the result.
(c) Compare the two intervals, part (a) and part (b). Which interval is wider? Will this always be the case? Explain.

:::