---
title: "Quiz 6"
# author: "Type Your Name Here"
# date: today

format: 
    pdf:
        pdf-engine: lualatex
        template-partials:
            - ../../doc-class.tex
        include-in-header:
        - text: |
            \AtBeginDocument{
                \renewcommand{\theexercise}{\arabic{exercise}}
            }
params:
    show_solution: false
filters:
    - ../../solution-filter.lua
crossref:
    chapters: false
---
A power company wants to use daily high temperature, $x$, to model daily peak power load, $y$, during the summer months when demand is greatest.

Although the company expects peak load to increase as the temperature increases, the rate of increase in $E(y)$ might not remain constant as $x$ increases. For example, a 1-unit increase in high temperature from 100 °F to 101 °F might result in a larger increase in power demand than would a 1-unit increase from 80 °F to 81 °F. Therefore, the company postulates that the model for $E(y)$ will include a second-order (quadratic) term and, possibly, a third-order (cubic) term.

A random sample of 25 summer days is selected and both the peak load (measured in megawatts) and high temperature (in degrees) recorded for each day. The R output is attached.



```{r}
#| echo: false
df <- data.frame(
    TEMP = c(
        94, 96, 95, 108, 67, 88, 89, 84, 90, # Column 1
        106, 67, 71, 100, 79, 97, 98, 87, # Column 2
        76, 68, 92, 100, 85, 89, 74, 86 # Column 3
    ),
    LOAD = c(
        136.0, 131.7, 140.7, 189.3, 96.5, 116.4, 118.5, 113.4, 132.0, # Col 1
        178.2, 101.6, 92.5, 151.9, 106.2, 153.2, 150.1, 114.7, # Col 2
        100.9, 96.3, 135.1, 143.6, 111.4, 116.5, 103.9, 105.1 # Col 3
    )
)
source("../../../anova_alt.R")
```

---

# 3RD ORDER MODEL

```{r}
third_oder_model <- lm(LOAD~TEMP+I(TEMP^2)+I(TEMP^3), data=df)
anova_alt(third_oder_model)
summary(third_oder_model)
```

---

# 2ND ORDER MODEL
```{r}
second_order_model <- lm(LOAD~TEMP+I(TEMP^2), data=df)
anova_alt(second_order_model)
summary(second_order_model)
```

{{< pagebreak >}} 

::: {#exr-}
Fit the third-order model, $E(y) = \beta_0 + \beta_1x + \beta_2x^2+\beta_3x^3$, to the data. Is there evidence that the cubic term, $\beta_3x^3$, contributes information for the prediction of peak power load? Test at $\alpha = .05$.
:::


::: {.answer}
We perform the t-test for the cubic term $H_0:\beta_3=0$. The p-value is 0.911. Since p-value > $\alpha$ = .05, we fail to reject null hypothesis. There is insufficient evidence of a third-order relationship between peak load and high temperature. Consequently, we will drop the cubic term, $\beta_3x^3$, from the model.
:::

::: {#exr-ref}
Fit the 2nd-order model, $E(y) = \beta_0 + \beta_1x + \beta_2x^2$, to the data.  Test the hypothesis that the power load increases at an increasing rate with temperature. In other words, is the curve concave upward? Use $\alpha = .05$.
:::


::: {.answer}
We fit the quadratic model. We test $H_0:\beta_2=0$. The two-tailed p-value is 0.000. Since the one-tailed p-value < $\alpha$ = .05, we reject null hypothesis. We conclude the peak power load increases at an increasing rate with temperature. In other words, the curve is concave upward.
:::


::: {#exr-}
Give the prediction equation for the second-order model, in @exr-ref. Are you satisfied with using this model to predict peak power loads?
:::


::: {.answer}
The prediction equation for the quadratic model is
$$
\hat y=385.0480-8.2925x+.0598x^2.
$$
Based on the F-test, the model is statistically useful.
:::