---
title: "HW 7"
# author: "Type Your Name Here"
# date: today
format: 
    pdf:
        pdf-engine: lualatex
        template-partials:
            - ../../doc-class.tex
        include-in-header:
        - ../../math.tex
        - text: |
            \AtBeginDocument{
                \renewcommand{\theexercise}{\arabic{exercise}}
            }
params:
    show_solution: false
filters:
    - ../../solution-filter.lua
crossref:
    chapters: false
---

 
::: {#exr-}
## Data set: `X.csv`

a. Fit the model $\Exp(y) = \beta_0 + \beta_1x$ to the data.
b. Calculate the residuals for the model.
c. Plot the residuals versus $x$. Do you detect any trends? If so, what does the pattern suggest about the model?
d. Plot the residuals for the first-order model versus $\hat y$. Do you detect any trends? If so, what does the pattern suggest about the model?
:::


::: {.answer}

```{r}
dat <- read.csv("X.csv")
fit <- lm(Y~., data=dat)
fit
```


```{r}
resid(fit)
```


```{r}
plot(dat$X, resid(fit))
```


```{r}
plot(fitted.values(fit), resid(fit))
```

```{r}
plot(fit, which=1)
```


a. From the R output, the Fitted model is $\hat y = −3.179 + 2.491x$.
c. From the residual plot, we see a quadratic U-shaped trend. For small and large values of $x$, the residuals are positive, while for intermediate values of $x$, the residuals are negative. This pattern indicates a lack-of-fit in the straight-line model and suggests to fit the quadratic model. To verify this, we would fit the quadratic model $\Exp(y) = \beta_0 + \beta_1x + \beta_2x^2$ and test for curvature, i.e., test $H_0 : \beta_2 = 0$.
d. Yes, there appears to be a trend. The assumption of constant variance appears to be satisfied. The pattern suggests adding curvature to the model.
:::



::: {#exr-}
## Data set: `R.csv`
Consider the dataset `R.csv` and build a 1st order MLR model with all variables.

a. Plot the model residuals against each of the four independent variables using R.
b. How would you advise the researchers to modify the model? Explain
::: 


::: {.answer}

```{r}
#| fig-show: hold
dat <- read.csv("R.csv")
fit <- lm(y ~ ., data = dat)
par(mfrow = c(2, 2))
plot(dat$x1, resid(fit))
abline(h = 0)
plot(dat$x2, resid(fit))
abline(h = 0)
plot(dat$x3, resid(fit))
abline(h = 0)
plot(dat$x4, resid(fit))
abline(h = 0)
par(mfrow = c(1, 1))
```
The plots of the residuals against both $x_1$ and $x_4$ appear to be curved. This implies that second-order terms for both $x_1$ and $x_4$ should be added to the model for a better fit.
:::